---
title: "Climate, Weather, and Stochastic Production Frontiers"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
params:
  climate_window_main: 20
  climate_window_tol: 30
  run_heavy_models: true
  run_year_fe_sfa: false
  cache_models: true
  results_dir: "results_sfa"
  project_setup_script: null
---

# Purpose

This notebook estimates a common family of models in which long-run climate affects the production frontier and weather anomalies affect realized output. It is designed around a national Penn World Table panel joined to annual temperature and precipitation data.

The main structural stochastic-frontier equation is

$$\ln y_{it} = \alpha_i + q_t + \beta \ln k_{it} + f(C^T_{it}, C^P_{it}) + v_{it} - u_{it},$$

where climate enters the frontier through $f(\cdot)$, while weather anomalies determine the one-sided component $u_{it}$.

The notebook estimates:

1.  OLS/fixed-effect diagnostics using `fixest`.
2.  A deterministic conditional-mean benchmark analogous to the climate-adaptation decomposition in Merel and Gammans.
3.  A Tol-core stochastic frontier benchmark: Greene-type country fixed effects, linear trend, exponential inefficiency, 30-year preceding climate, and absolute standardized weather anomalies.
4.  The same fixed-effect stochastic frontier with squared anomalies.
5.  The same fixed-effect stochastic frontier with asymmetric hot/cold/wet/dry anomalies.
6.  Greene (2005) `TFE` using the panel estimator in the current `sfa` package.
7.  Chen, Schmidt and Wang (2014) `TFE_WMLE`.
8.  `TRE_Z` with Mundlak/correlated-random-effects controls and weather determinants of inefficiency.
9.  Wang-Ho first-difference `FD`, which eliminates country effects while retaining a weather-dependent inefficiency scaling function.
10. Optional year-fixed-effect SFA variants.

The Tol comparison follows the core homogeneous specification in Tol (2021, arXiv:2102.13110): climate is the 30 years preceding year $t$; the baseline frontier contains temperature, temperature squared, precipitation, precipitation squared, and their interaction; the time control is a linear trend; inefficiency is exponential and depends on absolute standardized temperature and precipitation anomalies. Tol estimated a one-step True Fixed-Effect model following Greene (2005). He reports that a full set of year dummies caused convergence problems in the baseline stochastic-frontier likelihood. Tol also reports richer hot/poor heterogeneity specifications; these are not labeled as exact replications here because the present project does not currently supply the same World Bank income-classification variable.

Important: coefficients in the inefficiency equation are not directly comparable across all `sfa` model families because their link functions differ. The comparison section therefore reports both raw coefficients and model-specific scale ratios.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "dplyr", "tidyr", "purrr", "tibble", "ggplot2",
  "fixest", "e1071", "sfa", "digest"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages)) {
  stop(
    "Install missing packages before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)
library(fixest)
library(e1071)
library(sfa)

if (packageVersion("sfa") < package_version("1.1.3")) {
  stop(
    "This notebook expects sfa >= 1.1.3. In that version and later, ",
    "model_name='TFE' denotes Greene (2005) TFE and 'TFE_WMLE' denotes ",
    "Chen-Schmidt-Wang (2014) within MLE."
  )
}

MAIN_WINDOW <- params$climate_window_main
TOL_WINDOW  <- params$climate_window_tol
RUN_HEAVY   <- isTRUE(params$run_heavy_models)
RUN_YEAR_FE <- isTRUE(params$run_year_fe_sfa)
CACHE       <- isTRUE(params$cache_models)
RESULT_DIR  <- params$results_dir
ECON_YEAR_MIN <- 1940
CLIM_HISTORY_YEARS <- 30
  
# Optional project bootstrap. Set this parameter when knitting in a fresh R session,
# e.g. project_setup_script: "R/load_project_data.R".
if (!is.null(params$project_setup_script) &&
    length(params$project_setup_script) == 1L &&
    !is.na(params$project_setup_script) &&
    nzchar(params$project_setup_script)) {
  source(params$project_setup_script, local = globalenv())
}

required_project_objects <- c(
  "econ_data", "data_rr", "data_tm",
  "ECON_YEAR_MIN", "CLIM_HISTORY_YEARS", "lag_by_year"
)
missing_project_objects <- required_project_objects[
  !vapply(required_project_objects, exists, logical(1), inherits = TRUE)
]
if (length(missing_project_objects)) {
  stop(
    "Project objects are not loaded: ",
    paste(missing_project_objects, collapse = ", "),
    ". Source the project setup before knitting or set params$project_setup_script."
  )
}

if (CLIM_HISTORY_YEARS < TOL_WINDOW) {
  stop("CLIM_HISTORY_YEARS must be at least ", TOL_WINDOW,
       " to construct Tol-style preceding climate normals.")
}

dir.create(RESULT_DIR, showWarnings = FALSE, recursive = TRUE)

options(scipen = 5)
set.seed(12345)
```

# 2. Input data

This notebook uses the following objects from the existing project. They can already be loaded in the R session, or `project_setup_script` can point to a project bootstrap script that creates them:

- `econ_data`
- `data_rr`
- `data_tm`
- `ECON_YEAR_MIN`
- `CLIM_HISTORY_YEARS`
- `lag_by_year()`

The function below reproduces the current data construction. The economic sample begins at `ECON_YEAR_MIN`, while climate observations before that date are retained long enough to construct preceding climate moments.

```{r build-data}
build_dat <- function(w_rr, w_tm) {

  econ <- econ_data %>%
    filter(
      econ_source == "PWT110",
      year >= ECON_YEAR_MIN
    )

  data_rr %>%
    filter(
      window_rr == w_rr,
      weight == "area",
      climate_source == "era5",
      (weight_year == "2015" | weight_year == "un"),
      !is.na(RR)
    ) %>%
    select(-window_rr, -weight, -weight_year) %>%
    inner_join(
      data_tm %>%
        filter(
          window_tm == w_tm,
          weight == "area",
          climate_source == "era5",
          (weight_year == "2015" | weight_year == "un"),
          !is.na(TM)
        ) %>%
        select(-window_tm, -weight, -weight_year)
    ) %>%
    filter(year >= ECON_YEAR_MIN - CLIM_HISTORY_YEARS) %>%
    left_join(econ) %>%
    mutate(
      .has_level_outcome = !is.na(lgrp_pc_usd) |
        (!is.na(grp_pc_usd) & grp_pc_usd > 0)
    ) %>%
    group_by(GID_0) %>%
    # Keep countries with an estimable GDP-per-worker level and do not carry
    # climate observations past their final usable economic observation.
    filter(any(.has_level_outcome)) %>%
    filter(year <= max(year[.has_level_outcome])) %>%
    mutate(log_gdp_av = log(mean(grp_pc_usd, na.rm = TRUE))) %>%
    arrange(year, .by_group = TRUE) %>%
    mutate(
      mean_TM_lag = lag_by_year(mean_TM, year),
      mean_RR_lag = lag_by_year(mean_RR, year)
    ) %>%
    ungroup() %>%
    select(-.has_level_outcome)
}

# MAIN_WINDOW is used for the preferred specification.
df_raw <- build_dat(MAIN_WINDOW, MAIN_WINDOW)
```

## 2.1 Validate the panel

```{r validate-panel}
stopifnot(all(c("GID_0", "year", "TM", "RR") %in% names(df_raw)))

panel_gaps <- df_raw %>%
  arrange(GID_0, year) %>%
  group_by(GID_0) %>%
  summarise(
    n = n(),
    has_gap = any(diff(year) != 1),
    .groups = "drop"
  )

if (any(panel_gaps$has_gap)) {
  warning(
    sum(panel_gaps$has_gap),
    " countries have gaps in the climate panel. The rolling helper below requires ",
    "consecutive annual observations and returns NA where a full consecutive window is unavailable."
  )
}
```

# 3. Construct climate and weather variables

## 3.1 Helper for preceding rolling moments

Tol's climate at year $t$ is based on the 30 years preceding $t$, so the current weather realization must not enter its own climate normal. The helper below explicitly checks that the window contains consecutive years.

```{r rolling-helper}
previous_window_moment <- function(x, year, window, fun) {
  n <- length(x)
  out <- rep(NA_real_, n)

  for (i in seq_len(n)) {
    target_years <- seq.int(year[i] - window, year[i] - 1L)
    idx <- match(target_years, year)

    if (!anyNA(idx)) {
      vals <- x[idx]
      if (all(is.finite(vals))) {
        out[i] <- fun(vals)
      }
    }
  }

  out
}

make_preceding_moments <- function(data, window, prefix) {
  data %>%
    arrange(GID_0, year) %>%
    group_by(GID_0) %>%
    mutate(
      "{prefix}_TM_mean" := previous_window_moment(TM, year, window, mean),
      "{prefix}_TM_sd"   := previous_window_moment(TM, year, window, stats::sd),
      "{prefix}_RR_mean" := previous_window_moment(RR, year, window, mean),
      "{prefix}_RR_sd"   := previous_window_moment(RR, year, window, stats::sd)
    ) %>%
    ungroup()
}
```

## 3.2 Main and Tol climate moments

For the main specification, the existing lagged rolling means are retained where available. Standard deviations use the project's `sd_TM_all` and `sd_RR_all`, matching the current specification. A separate set of preceding 30-year mean and SD variables is computed for the Tol comparison.

```{r construct-moments}
df <- df_raw %>%
  make_preceding_moments(window = TOL_WINDOW, prefix = "tol")

if (!all(c("mean_TM_lag", "mean_RR_lag", "sd_TM_all", "sd_RR_all") %in% names(df))) {
  stop(
    "The preferred specification expects mean_TM_lag, mean_RR_lag, sd_TM_all and sd_RR_all. ",
    "Adjust this section if the project uses different names."
  )
}

df <- df %>%
  mutate(
    # Correct conversion from per-capita variables to per-worker variables.
    # If lgrp_pc_usd = log(GDP/pop), then log(GDP/worker) = lgrp_pc_usd - log(worker/pop).
    y = lgrp_pc_usd - log(share_emp_pop),
    k = log(k_pc_usd) - log(share_emp_pop),

    # Preferred climate frontier.
    clim_T = mean_TM_lag,
    clim_P = mean_RR_lag,
    clim_T2 = clim_T^2,
    clim_P2 = clim_P^2,
    clim_TP = clim_T * clim_P,

    # Preferred weather standardization.
    zT = (TM - clim_T) / sd_TM_all,
    zP = (RR - clim_P) / sd_RR_all,
    zT2 = zT^2,
    zP2 = zP^2,
    abs_zT = abs(zT),
    abs_zP = abs(zP),
    hot = pmax(zT, 0),
    cold = pmax(-zT, 0),
    wet = pmax(zP, 0),
    dry = pmax(-zP, 0),

    # Tol-exact climate/weather construction: preceding 30-year mean and SD.
    tol_clim_T = tol_TM_mean,
    tol_clim_P = tol_RR_mean,
    tol_clim_T2 = tol_clim_T^2,
    tol_clim_P2 = tol_clim_P^2,
    tol_clim_TP = tol_clim_T * tol_clim_P,
    tol_zT = (TM - tol_clim_T) / tol_TM_sd,
    tol_zP = (RR - tol_clim_P) / tol_RR_sd,
    tol_abs_zT = abs(tol_zT),
    tol_abs_zP = abs(tol_zP),
    tol_zT2 = tol_zT^2,
    tol_zP2 = tol_zP^2,

    trend = year - min(year, na.rm = TRUE)
  )
```

Check that the two ways of constructing GDP per worker agree when the level variable is available:

```{r check-output-transform}
if ("grp_pc_usd" %in% names(df)) {
  transform_check <- df %>%
    filter(
      is.finite(y),
      is.finite(grp_pc_usd),
      is.finite(share_emp_pop),
      grp_pc_usd > 0,
      share_emp_pop > 0
    ) %>%
    summarise(
      max_abs_difference = max(
        abs(y - log(grp_pc_usd / share_emp_pop)),
        na.rm = TRUE
      )
    )

  print(transform_check)
}
```

# 4. Common estimation sample

Using a common sample is essential for meaningful likelihood and coefficient comparisons. The common sample below contains all variables needed by the core models.

```{r common-sample}
common_vars <- c(
  "GID_0", "year", "y", "k", "TM", "RR", "trend",
  "clim_T", "clim_T2", "clim_P", "clim_P2", "clim_TP",
  "zT", "zP", "zT2", "zP2", "abs_zT", "abs_zP",
  "hot", "cold", "wet", "dry",
  "tol_clim_T", "tol_clim_T2", "tol_clim_P", "tol_clim_P2", "tol_clim_TP",
  "tol_zT", "tol_zP", "tol_abs_zT", "tol_abs_zP", "tol_zT2", "tol_zP2"
)

missing_common <- setdiff(common_vars, names(df))
if (length(missing_common)) {
  stop("Missing variables: ", paste(missing_common, collapse = ", "))
}

df_sfa <- df %>%
  filter(if_all(all_of(common_vars), ~ !is.na(.x))) %>%
  filter(
    if_all(
      all_of(setdiff(common_vars, "GID_0")),
      ~ is.finite(as.numeric(.x))
    )
  ) %>%
  group_by(GID_0) %>%
  filter(n() >= 3) %>%
  ungroup() %>%
  arrange(GID_0, year) %>%
  mutate(
    country_fe = droplevels(factor(GID_0)),
    year_fe = droplevels(factor(year))
  )

sample_summary <- df_sfa %>%
  summarise(
    observations = n(),
    countries = n_distinct(GID_0),
    first_year = min(year),
    last_year = max(year),
    min_T_per_country = min(table(GID_0)),
    max_T_per_country = max(table(GID_0))
  )

print(sample_summary)
```

# 5. OLS and fixed-effect diagnostics

These regressions do not estimate a stochastic frontier. They are diagnostics for residual asymmetry and benchmarks for the frontier coefficients.

```{r ols-diagnostics}
ols_pooled <- feols(
  y ~ k + clim_T + clim_T2 + clim_P + clim_P2 + clim_TP,
  data = df_sfa
)

ols_country_fe <- feols(
  y ~ k + clim_T + clim_T2 + clim_P + clim_P2 + clim_TP | GID_0,
  data = df_sfa
)

ols_country_year_fe <- feols(
  y ~ k + clim_T + clim_T2 + clim_P + clim_P2 + clim_TP | GID_0 + year,
  cluster = ~GID_0,
  data = df_sfa
)

residual_diagnostics <- tibble(
  model = c("Pooled OLS", "Country FE", "Country + year FE"),
  skewness = c(
    e1071::skewness(resid(ols_pooled), type = 2),
    e1071::skewness(resid(ols_country_fe), type = 2),
    e1071::skewness(resid(ols_country_year_fe), type = 2)
  ),
  sd_residual = c(
    sd(resid(ols_pooled)),
    sd(resid(ols_country_fe)),
    sd(resid(ols_country_year_fe))
  )
)

print(residual_diagnostics)
```

Negative residual skewness is consistent with a production frontier whose composite residual is $v-u$, but it is not sufficient to identify weather as the source of the one-sided component.

# 6. Deterministic conditional-mean benchmarks

## 6.1 Climate frontier plus squared weather anomaly in the conditional mean

This model asks whether unusual weather predicts average output directly, rather than the distribution of a latent one-sided output gap.

```{r deterministic-benchmark}
mean_quad_fe <- feols(
  y ~
    k + clim_T + clim_T2 + clim_P + clim_P2 + clim_TP +
    zT2 + zP2 |
    GID_0 + year,
  cluster = ~GID_0,
  data = df_sfa
)

summary(mean_quad_fe)
```

## 6.2 Merel-Gammans-style adaptation benchmark

This is a conceptual analogue rather than a literal replication of their application. Actual weather enters the conditional mean, while squared deviations from the climate normal capture an additional unfamiliar-weather penalty.

```{r merel-benchmark}
merel_like <- feols(
  y ~
    k +
    TM + I(TM^2) +
    RR + I(RR^2) +
    zT2 + zP2 |
    GID_0 + year,
  cluster = ~GID_0,
  data = df_sfa
)

summary(merel_like)
```

# 7. Helpers for stochastic-frontier estimation

The models can be slow. Each successful fit is cached as an `.rds` file so that knitting the notebook again does not re-estimate unchanged models.

```{r sfa-helpers}
is_sfa_fit <- function(x) inherits(x, "sfareg")
`%||%` <- function(x, y) if (is.null(x)) y else x

MODEL_CACHE_TAG <- substr(
  digest::digest(
    list(
      data = df_sfa[, common_vars, drop = FALSE],
      main_window = MAIN_WINDOW,
      tol_window = TOL_WINDOW,
      sfa_version = as.character(packageVersion("sfa"))
    ),
    algo = "xxhash64"
  ),
  1, 12
)

fit_cached <- function(name, fit_fun, overwrite = FALSE) {
  path <- file.path(RESULT_DIR, paste0(name, "_", MODEL_CACHE_TAG, ".rds"))

  if (CACHE && file.exists(path) && !overwrite) {
    message("Loading cached model: ", name)
    return(readRDS(path))
  }

  message("Estimating model: ", name)
  t0 <- Sys.time()

  fit <- tryCatch(
    fit_fun(),
    error = function(e) {
      structure(
        list(
          model_name = name,
          error = conditionMessage(e),
          elapsed_seconds = as.numeric(difftime(Sys.time(), t0, units = "secs"))
        ),
        class = "sfa_fit_error"
      )
    }
  )

  # Cache successful model objects only. A failed optimization should be retried
  # after code/data changes rather than silently reloaded forever.
  if (CACHE && is_sfa_fit(fit)) saveRDS(fit, path)
  fit
}

safe_logLik <- function(fit) {
  if (!is_sfa_fit(fit)) return(NA_real_)
  tryCatch(as.numeric(logLik(fit)), error = function(e) NA_real_)
}

safe_AIC <- function(fit) {
  if (!is_sfa_fit(fit)) return(NA_real_)
  tryCatch(AIC(fit), error = function(e) NA_real_)
}

safe_BIC <- function(fit) {
  if (!is_sfa_fit(fit)) return(NA_real_)
  tryCatch(BIC(fit), error = function(e) NA_real_)
}

safe_nobs <- function(fit) {
  if (!is_sfa_fit(fit)) return(NA_integer_)
  tryCatch(nobs(fit), error = function(e) nrow(fit$data))
}

tidy_sfa <- function(fit, label) {
  if (!is_sfa_fit(fit)) {
    return(tibble(
      model = label,
      term = NA_character_,
      estimate = NA_real_,
      std_error = NA_real_,
      t_value = NA_real_,
      error = fit$error %||% "unknown error"
    ))
  }

  raw <- as.matrix(fit$out)

  # sfa versions have exposed fit$out in either p x 3 or 3 x p orientation.
  # Normalize to p x 3 without relying on undocumented orientation.
  if (
    nrow(raw) == 3L &&
    !is.null(rownames(raw)) &&
    all(c("par", "st_err", "t-val") %in% rownames(raw))
  ) {
    raw <- t(raw)
  }

  if (ncol(raw) < 3L) {
    stop("Unexpected fit$out structure for model ", label)
  }

  terms <- rownames(raw)
  if (is.null(terms)) terms <- paste0("par_", seq_len(nrow(raw)))

  tibble(
    model = label,
    term = terms,
    estimate = as.numeric(raw[, 1]),
    std_error = as.numeric(raw[, 2]),
    t_value = as.numeric(raw[, 3]),
    error = NA_character_
  )
}
```

# 8. Tol-core direct methodological comparison

Tol's equation is

$$\ln y_{it} = \beta \ln k_{it} + f(\bar T_{it}, \bar R_{it}) + \mu_i + t + v_{it} - u_{it},$$

where $\bar T$ and $\bar R$ are averages over the 30 years preceding $t$, and

$$u_{it}\sim \mathrm{Exponential}(\lambda_{it}),$$

with the inefficiency scale depending on absolute standardized weather anomalies. The model below uses `sfm(..., model_name = "NE_Z")` and explicit country dummies. This is the closest current `sfa` implementation to Tol's core one-step Greene TFE with exponential, heteroskedastic inefficiency.

It is a methodological comparison, not a numerical or exact preferred-column replication: the present data use PWT 11.0, ERA5, the project's spatial weights, and a different sample. Tol's richer preferred specifications also interact climate/weather terms with hot/poor-country indicators.

```{r tol-model}
models <- list()

models$tol_abs_fe <- fit_cached(
  "tol_abs_fe",
  function() {
    sfa::sfm(
      y ~
        k +
        tol_clim_T + tol_clim_T2 +
        tol_clim_P + tol_clim_P2 +
        tol_clim_TP +
        trend +
        country_fe |
        tol_abs_zT + tol_abs_zP,
      model_name = "NE_Z",
      data = df_sfa
    )
  }
)

if (is_sfa_fit(models$tol_abs_fe)) summary(models$tol_abs_fe)
```

# 9. Preferred squared-anomaly fixed-effect SFA

This changes the weather metric while holding the country-FE treatment, exponential inefficiency distribution, frontier, and trend fixed.

```{r quad-fe-model}
models$quad_fe <- fit_cached(
  "quad_fe",
  function() {
    sfa::sfm(
      y ~
        k +
        clim_T + clim_T2 +
        clim_P + clim_P2 +
        clim_TP +
        trend +
        country_fe |
        zT2 + zP2,
      model_name = "NE_Z",
      data = df_sfa
    )
  }
)

if (is_sfa_fit(models$quad_fe)) summary(models$quad_fe)
```

# 10. Asymmetric weather fixed-effect SFA

```{r asym-fe-model}
models$asym_fe <- fit_cached(
  "asym_fe",
  function() {
    sfa::sfm(
      y ~
        k +
        clim_T + clim_T2 +
        clim_P + clim_P2 +
        clim_TP +
        trend +
        country_fe |
        hot + cold + wet + dry,
      model_name = "NE_Z",
      data = df_sfa
    )
  }
)

if (is_sfa_fit(models$asym_fe)) summary(models$asym_fe)
```

# 11. Greene TFE and within-MLE fixed effects

These models are useful for assessing the treatment of persistent country heterogeneity. In the current `sfa` package, `TFE` and `TFE_WMLE` do not provide the same weather-dependent `_Z` inefficiency equation used above. They therefore test the frontier and fixed-effect estimator, not the complete weather mechanism.

## 11.1 Greene (2005) TFE

```{r greene-tfe}
if (RUN_HEAVY) {
  models$tfe_greene <- fit_cached(
    "tfe_greene",
    function() {
      sfa::psfm(
        y ~
          k +
          clim_T + clim_T2 +
          clim_P + clim_P2 +
          clim_TP +
          trend,
        model_name = "TFE",
        data = df_sfa,
        individual = "GID_0",
        time = "year"
      )
    }
  )

  if (is_sfa_fit(models$tfe_greene)) summary(models$tfe_greene)
}
```

A warning that `lambda` reaches its upper bound should be treated as a boundary/degeneracy warning, not as evidence of precisely estimated near-zero symmetric noise. The current package explicitly recommends `TFE_WMLE` when Greene TFE runs to this boundary.

## 11.2 Chen-Schmidt-Wang within MLE

```{r tfe-wmle}
if (RUN_HEAVY) {
  models$tfe_wmle <- fit_cached(
    "tfe_wmle",
    function() {
      sfa::psfm(
        y ~
          k +
          clim_T + clim_T2 +
          clim_P + clim_P2 +
          clim_TP +
          trend,
        model_name = "TFE_WMLE",
        data = df_sfa,
        individual = "GID_0",
        time = "year"
      )
    }
  )

  if (is_sfa_fit(models$tfe_wmle)) summary(models$tfe_wmle)
}
```

# 12. Mundlak correlated random effects for `TRE_Z`

`TRE_Z` allows weather determinants of the time-varying one-sided component, but country heterogeneity is random rather than fixed. Mundlak controls are added to allow the persistent country effect to correlate with the observed time-varying frontier regressors.

The country means are constructed on the final common estimation sample.

```{r mundlak-data}
df_sfa <- df_sfa %>%
  group_by(GID_0) %>%
  mutate(
    k_bar = mean(k),
    clim_T_bar = mean(clim_T),
    clim_T2_bar = mean(clim_T2),
    clim_P_bar = mean(clim_P),
    clim_P2_bar = mean(clim_P2),
    clim_TP_bar = mean(clim_TP),

    tol_clim_T_bar = mean(tol_clim_T),
    tol_clim_T2_bar = mean(tol_clim_T2),
    tol_clim_P_bar = mean(tol_clim_P),
    tol_clim_P2_bar = mean(tol_clim_P2),
    tol_clim_TP_bar = mean(tol_clim_TP)
  ) %>%
  ungroup()
```

## 12.1 Preferred squared anomalies

```{r tre-mundlak-quad}
if (RUN_HEAVY) {
  models$tre_mundlak_quad <- fit_cached(
    "tre_mundlak_quad",
    function() {
      sfa::psfm(
        y ~
          k +
          clim_T + clim_T2 +
          clim_P + clim_P2 +
          clim_TP +
          trend +
          k_bar +
          clim_T_bar + clim_T2_bar +
          clim_P_bar + clim_P2_bar +
          clim_TP_bar |
          zT2 + zP2,
        model_name = "TRE_Z",
        data = df_sfa,
        individual = "GID_0",
        time = "year",
        collinear_action = "start_only"
      )
    }
  )

  if (is_sfa_fit(models$tre_mundlak_quad)) summary(models$tre_mundlak_quad)
}
```

## 12.2 Tol-type absolute anomalies

```{r tre-mundlak-abs}
if (RUN_HEAVY) {
  models$tre_mundlak_abs <- fit_cached(
    "tre_mundlak_abs",
    function() {
      sfa::psfm(
        y ~
          k +
          tol_clim_T + tol_clim_T2 +
          tol_clim_P + tol_clim_P2 +
          tol_clim_TP +
          trend +
          k_bar +
          tol_clim_T_bar + tol_clim_T2_bar +
          tol_clim_P_bar + tol_clim_P2_bar +
          tol_clim_TP_bar |
          tol_abs_zT + tol_abs_zP,
        model_name = "TRE_Z",
        data = df_sfa,
        individual = "GID_0",
        time = "year",
        collinear_action = "start_only"
      )
    }
  )

  if (is_sfa_fit(models$tre_mundlak_abs)) summary(models$tre_mundlak_abs)
}
```

For `TRE_Z`, the package parameterizes the inefficiency variance as

$$\sigma_{u,it}^2 = \exp(z_{it}'\delta),$$

so the standard deviation is $\sigma_{u,it}=\exp(z_{it}'\delta/2)$.

# 13. Wang-Ho first-difference estimator

The `FD` estimator removes time-invariant country effects by first differencing the frontier while modeling time-varying inefficiency as a scaling function. It therefore provides a useful fixed-effect robustness check without manually demeaning the composite SFA error.

In the current `sfa` implementation, the scaling function is

$$h_{it}=\exp(z_{it}'\delta).$$

## 13.1 Squared anomalies

```{r fd-quad}
if (RUN_HEAVY) {
  models$fd_quad <- fit_cached(
    "fd_quad",
    function() {
      sfa::psfm(
        y ~
          k +
          clim_T + clim_T2 +
          clim_P + clim_P2 +
          clim_TP +
          trend |
          zT2 + zP2,
        model_name = "FD",
        data = df_sfa,
        individual = "GID_0",
        time = "year"
      )
    }
  )

  if (is_sfa_fit(models$fd_quad)) summary(models$fd_quad)
}
```

## 13.2 Absolute anomalies

```{r fd-abs}
if (RUN_HEAVY) {
  models$fd_abs <- fit_cached(
    "fd_abs",
    function() {
      sfa::psfm(
        y ~
          k +
          tol_clim_T + tol_clim_T2 +
          tol_clim_P + tol_clim_P2 +
          tol_clim_TP +
          trend |
          tol_abs_zT + tol_abs_zP,
        model_name = "FD",
        data = df_sfa,
        individual = "GID_0",
        time = "year"
      )
    }
  )

  if (is_sfa_fit(models$fd_abs)) summary(models$fd_abs)
}
```

Do not interpret the FD model as literally the same stochastic process as the levels model. Tol also notes that differencing changes the distributional structure of the one-sided error. Here FD is a robustness estimator for persistent country heterogeneity.

# 14. Optional non-parametric year effects in SFA

Tol used a linear trend because year dummies made the baseline maximum-likelihood problem computationally cumbersome. The following models are therefore off by default. Enable them using `params$run_year_fe_sfa: true`.

```{r optional-year-fe}
if (RUN_YEAR_FE) {
  models$quad_fe_year <- fit_cached(
    "quad_fe_year",
    function() {
      sfa::sfm(
        y ~
          k +
          clim_T + clim_T2 +
          clim_P + clim_P2 +
          clim_TP +
          country_fe + year_fe |
          zT2 + zP2,
        model_name = "NE_Z",
        data = df_sfa
      )
    }
  )

  models$tre_mundlak_quad_year <- fit_cached(
    "tre_mundlak_quad_year",
    function() {
      sfa::psfm(
        y ~
          k +
          clim_T + clim_T2 +
          clim_P + clim_P2 +
          clim_TP +
          year_fe +
          k_bar +
          clim_T_bar + clim_T2_bar +
          clim_P_bar + clim_P2_bar +
          clim_TP_bar |
          zT2 + zP2,
        model_name = "TRE_Z",
        data = df_sfa,
        individual = "GID_0",
        time = "year",
        collinear_action = "start_only"
      )
    }
  )
}
```

# 15. Model-status table

```{r model-status}
model_metadata <- tribble(
  ~model, ~country_effect, ~time_control, ~u_structure, ~weather_metric, ~role,
  "tol_abs_fe", "Explicit country FE", "Linear trend", "Exponential, heteroskedastic", "Absolute z, 30y preceding", "Closest Tol comparison",
  "quad_fe", "Explicit country FE", "Linear trend", "Exponential, heteroskedastic", "Squared z", "Preferred metric, same FE likelihood",
  "asym_fe", "Explicit country FE", "Linear trend", "Exponential, heteroskedastic", "Hot/cold/wet/dry", "Asymmetry robustness",
  "tfe_greene", "Greene TFE", "Linear trend", "Half-normal TFE", "No weather Z equation", "Country-FE frontier robustness",
  "tfe_wmle", "Within MLE", "Linear trend", "Within SFA", "No weather Z equation", "Demeaning-consistent SFA robustness",
  "tre_mundlak_quad", "TRE + Mundlak", "Linear trend", "TRE_Z", "Squared z", "Practical main alternative",
  "tre_mundlak_abs", "TRE + Mundlak", "Linear trend", "TRE_Z", "Absolute z, 30y preceding", "Tol metric under correlated RE",
  "fd_quad", "First difference", "Differenced trend", "Wang-Ho scaling", "Squared z", "Fixed-effect robustness",
  "fd_abs", "First difference", "Differenced trend", "Wang-Ho scaling", "Absolute z, 30y preceding", "Tol metric under FD"
)

model_status <- model_metadata %>%
  mutate(
    object = map(model, ~ models[[.x]]),
    estimated = map_lgl(object, is_sfa_fit),
    error = map_chr(object, ~ {
      if (is.null(.x)) return("not run")
      if (inherits(.x, "sfa_fit_error")) return(.x$error)
      ""
    }),
    nobs = map_int(object, ~ {
      ans <- safe_nobs(.x)
      ifelse(is.na(ans), NA_integer_, as.integer(ans))
    }),
    logLik = map_dbl(object, safe_logLik),
    AIC = map_dbl(object, safe_AIC),
    BIC = map_dbl(object, safe_BIC)
  ) %>%
  select(-object)

print(model_status)
```

# 16. Coefficient comparison

```{r coefficient-comparison}
coef_long <- imap_dfr(
  models,
  ~ tidy_sfa(.x, .y)
)

frontier_terms <- c(
  "k",
  "clim_T", "clim_T2", "clim_P", "clim_P2", "clim_TP",
  "tol_clim_T", "tol_clim_T2", "tol_clim_P", "tol_clim_P2", "tol_clim_TP",
  "trend"
)

frontier_comparison <- coef_long %>%
  filter(term %in% frontier_terms) %>%
  select(model, term, estimate, std_error, t_value) %>%
  arrange(term, model)

print(frontier_comparison)

weather_terms <- c(
  "zT2", "zP2", "abs_zT", "abs_zP",
  "tol_zT2", "tol_zP2", "tol_abs_zT", "tol_abs_zP",
  "hot", "cold", "wet", "dry"
)

weather_comparison_raw <- coef_long %>%
  filter(term %in% weather_terms) %>%
  select(model, term, estimate, std_error, t_value) %>%
  arrange(term, model)

print(weather_comparison_raw)
```

# 17. Convert weather coefficients to comparable scale ratios

A raw coefficient on a weather determinant has a different meaning across model families.

For `sfm(..., model_name="NE_Z")`:

$$\sigma_u(z)=\exp(z'\delta),$$

so a one-unit increase in a weather determinant multiplies the inefficiency scale by $\exp(\delta)$.

For `psfm(..., model_name="TRE_Z")`:

$$\sigma_u^2(z)=\exp(z'\delta),$$

so the standard deviation multiplies by $\exp(\delta/2)$.

For `FD`:

$$h(z)=\exp(z'\delta),$$

so a one-unit increase multiplies the inefficiency scaling factor by $\exp(\delta)$.

```{r weather-scale-ratios}
model_family <- c(
  tol_abs_fe = "NE_Z",
  quad_fe = "NE_Z",
  asym_fe = "NE_Z",
  tre_mundlak_quad = "TRE_Z",
  tre_mundlak_abs = "TRE_Z",
  fd_quad = "FD",
  fd_abs = "FD",
  quad_fe_year = "NE_Z",
  tre_mundlak_quad_year = "TRE_Z"
)

weather_scale_ratios <- weather_comparison_raw %>%
  mutate(
    family = unname(model_family[model]),
    scale_ratio_1unit = case_when(
      family == "NE_Z" ~ exp(estimate),
      family == "TRE_Z" ~ exp(estimate / 2),
      family == "FD" ~ exp(estimate),
      TRUE ~ NA_real_
    )
  )

print(weather_scale_ratios)
```

For `zT2`, moving from normal weather $z=0$ to a one-standard-deviation anomaly $|z|=1$ changes `zT2` by one unit. Thus `scale_ratio_1unit` has a direct interpretation as the multiplicative change in the relevant inefficiency scale for a one-SD temperature anomaly, holding other determinants fixed. The same logic applies to precipitation.

For absolute anomalies, a change from $|z|=0$ to $|z|=1$ is also a one-unit change.

# 18. Compare model-implied inefficiency/efficiency where available

Different estimators return different efficiency objects. The helper below extracts a common observation-level quantity where possible.

```{r efficiency-extraction}
extract_efficiency <- function(fit, model_name) {
  if (!is_sfa_fit(fit)) return(NULL)

  n <- nrow(fit$data)

  if (!is.null(fit$exp_u_hat) && length(fit$exp_u_hat) == n) {
    return(tibble(model = model_name, efficiency = as.numeric(fit$exp_u_hat)))
  }

  if (!is.null(fit$U) && length(fit$U) == n) {
    return(tibble(model = model_name, efficiency = as.numeric(fit$U)))
  }

  NULL
}

efficiency_long <- imap_dfr(models, extract_efficiency)

if (nrow(efficiency_long)) {
  efficiency_summary <- efficiency_long %>%
    group_by(model) %>%
    summarise(
      mean_efficiency = mean(efficiency, na.rm = TRUE),
      median_efficiency = median(efficiency, na.rm = TRUE),
      p05 = quantile(efficiency, 0.05, na.rm = TRUE),
      p95 = quantile(efficiency, 0.95, na.rm = TRUE),
      .groups = "drop"
    )

  print(efficiency_summary)
}
```

Do not compare mean efficiency mechanically across models with different distributions and decompositions. The more informative comparison is whether weather coefficients have stable signs and whether predicted weather-related scale changes are similar.

# 19. Frontier-response comparison

This section compares the fitted climate response holding precipitation or temperature at a reference value. Only models containing the main quadratic climate frontier are used.

```{r frontier-response-functions}
get_estimate <- function(fit, term) {
  if (!is_sfa_fit(fit)) return(NA_real_)
  tab <- tidy_sfa(fit, "tmp")
  hit <- tab$estimate[tab$term == term]
  if (length(hit) != 1) return(NA_real_)
  hit
}

main_frontier_models <- intersect(
  c("quad_fe", "asym_fe", "tfe_greene", "tfe_wmle", "tre_mundlak_quad", "fd_quad"),
  names(models)
)

reference_P <- median(df_sfa$clim_P, na.rm = TRUE)
T_grid <- seq(
  quantile(df_sfa$clim_T, 0.02, na.rm = TRUE),
  quantile(df_sfa$clim_T, 0.98, na.rm = TRUE),
  length.out = 100
)

frontier_T_curves <- map_dfr(main_frontier_models, function(mn) {
  fit <- models[[mn]]
  if (!is_sfa_fit(fit)) return(NULL)

  bT  <- get_estimate(fit, "clim_T")
  bT2 <- get_estimate(fit, "clim_T2")
  bTP <- get_estimate(fit, "clim_TP")

  if (anyNA(c(bT, bT2, bTP))) return(NULL)

  value <- bT * T_grid + bT2 * T_grid^2 + bTP * T_grid * reference_P
  value <- value - value[which.min(abs(T_grid - median(df_sfa$clim_T, na.rm = TRUE)))]

  tibble(model = mn, climate_T = T_grid, relative_log_frontier = value)
})

if (nrow(frontier_T_curves)) {
  ggplot(frontier_T_curves, aes(climate_T, relative_log_frontier, linetype = model)) +
    geom_line(linewidth = 0.8) +
    labs(
      x = "Long-run temperature",
      y = "Relative log production frontier",
      title = "Climate-frontier response across SFA estimators",
      subtitle = paste0("Precipitation fixed at sample median = ", round(reference_P, 3))
    ) +
    theme_minimal()
}
```

# 20. Final comparison table

The table below is the main output for methodological comparison.

```{r final-comparison}
final_comparison <- model_status %>%
  mutate(
    interpretation = case_when(
      model == "tol_abs_fe" ~ "Direct Tol-method benchmark: climate frontier + absolute standardized weather in exponential inefficiency.",
      model == "quad_fe" ~ "Same likelihood and FE treatment as Tol benchmark, replacing |z| with z^2.",
      model == "asym_fe" ~ "Allows hot/cold and wet/dry shocks to affect inefficiency differently.",
      model == "tfe_greene" ~ "Greene TFE estimated by panel routine; tests frontier under direct FE, but no weather-Z inefficiency equation.",
      model == "tfe_wmle" ~ "Within-MLE fixed-effects SFA; principled alternative to manual demeaning, but no weather-Z equation.",
      model == "tre_mundlak_quad" ~ "Weather-dependent inefficiency with correlated-random-effects approximation to country FE.",
      model == "tre_mundlak_abs" ~ "Tol weather metric under TRE_Z + Mundlak rather than fixed effects.",
      model == "fd_quad" ~ "Country effects removed by Wang-Ho transformation; weather scales inefficiency through z^2.",
      model == "fd_abs" ~ "Country effects removed by Wang-Ho transformation; Tol-type |z| weather scaling.",
      TRUE ~ ""
    )
  )

print(final_comparison)
```

# 21. Decision rules for the empirical paper

Use the following hierarchy when interpreting the estimates.

1.  **Tol comparability.** `tol_abs_fe` is the closest methodological benchmark. Differences from Tol should first be attributed to data, sample, climate source/weighting, and then to coefficient heterogeneity not included in the homogeneous benchmark.
2.  **Shock metric.** Compare `tol_abs_fe` with `quad_fe`. This isolates absolute versus squared standardized anomalies while holding the stochastic distribution and fixed-effect treatment constant.
3.  **Shock asymmetry.** Compare `quad_fe` with `asym_fe`. A large likelihood improvement or materially different hot/cold and wet/dry effects argues against the symmetric quadratic restriction.
4.  **Country heterogeneity.** Compare frontier coefficients from explicit-FE, `TFE`, `TFE_WMLE`, `TRE_Z + Mundlak`, and `FD`. Stable climate coefficients are more informative than raw likelihood rankings because these estimators impose different error structures.
5.  **Weather robustness.** Compare model-specific one-SD weather scale ratios from explicit-FE `NE_Z`, `TRE_Z + Mundlak`, and `FD`. Focus on sign, magnitude, and stability rather than raw delta coefficients.
6.  **Time controls.** Only after the baseline models converge reliably, enable the optional year-FE SFA variants. Tol's own baseline used a linear trend because year dummies caused convergence problems.
7.  **Do not manually demean and then call an ordinary SFA likelihood.** Demeaning transforms $u_{it}$ into $u_{it}-\bar u_i$, which is no longer one-sided. Use `TFE_WMLE` or `FD` when the purpose is to remove fixed effects by transformation.

# 22. Save outputs

```{r save-results}
write.csv(residual_diagnostics, file.path(RESULT_DIR, "residual_diagnostics.csv"), row.names = FALSE)
write.csv(model_status, file.path(RESULT_DIR, "model_status.csv"), row.names = FALSE)
write.csv(frontier_comparison, file.path(RESULT_DIR, "frontier_coefficients.csv"), row.names = FALSE)
write.csv(weather_comparison_raw, file.path(RESULT_DIR, "weather_coefficients_raw.csv"), row.names = FALSE)
write.csv(weather_scale_ratios, file.path(RESULT_DIR, "weather_scale_ratios.csv"), row.names = FALSE)
write.csv(final_comparison, file.path(RESULT_DIR, "final_model_comparison.csv"), row.names = FALSE)

saveRDS(
  list(
    models = models,
    sample_summary = sample_summary,
    residual_diagnostics = residual_diagnostics,
    final_comparison = final_comparison
  ),
  file.path(RESULT_DIR, "all_sfa_results.rds")
)
```

# 23. Session information

```{r session-info}
sessionInfo()
```

# References and implementation notes

- Tol, R. S. J. (2021), *The economic impact of weather and climate*, arXiv:2102.13110. The baseline uses the 30 years preceding year $t$, country fixed effects, a linear trend, an exponential inefficiency distribution, and absolute standardized weather anomalies. Squared anomalies are a robustness specification.
- Greene, W. (2005), *Fixed and random effects in stochastic frontier models*, Journal of Productivity Analysis 23:7-32.
- Chen, Y.-Y., Schmidt, P., and Wang, H.-J. (2014), within-MLE fixed-effects stochastic-frontier estimator.
- Wang-Ho first-difference SFA provides a transformation-based treatment of individual effects with a time-varying inefficiency scaling function.
- Current `sfa` package interface checked against the 2026 package documentation/source. `TFE` denotes Greene TFE in versions \>= 1.1.3; `TFE_WMLE` denotes Chen-Schmidt-Wang. `TRE_Z` models the variance of the one-sided component; `FD` supports a `| z` scaling equation.
