---
title: "Additional_analyses"
output: html_document
date: "2026-09-29"
---


## Temperature and precipitation level bins with signed deviations

This specification adds nonparametric bins for annual temperature and
precipitation levels to the joint signed-deviation model. Temperature bins use
5-degree Celsius intervals and precipitation bins use 0.5-metre intervals. The
most populated central intervals (10--15 degrees Celsius and 0.5--1 metre) are
the level-bin references; the signed-deviation reference remains -0.5 to 0.5
within-region standard deviations. Consequently, every plotted coefficient is
conditional on the common BHM climate controls and on the other three bin sets.

```{r test-level-and-signed-bins, fig.width=8, fig.height=5, fig.show='hold', out.width='50%'}
model_bins_level_signed <- paste(
  "+ i(TM_bin_level)",
  "+ i(RR_bin_level)",
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
)
f_bins_level_signed <- make_formula(model_bins_level_signed)
m_bins_level_signed <- fit_climate_model(model_bins_level_signed)

iplot(
  m_bins_level_signed,
  i.select = 1,
  main = "Annual temperature bins",
  xlab = "Annual mean temperature (degrees Celsius)"
)
iplot(
  m_bins_level_signed,
  i.select = 2,
  main = "Annual precipitation bins",
  xlab = "Annual precipitation (metres)"
)
iplot(
  m_bins_level_signed,
  i.select = 3,
  main = "Signed temperature-deviation bins",
  xlab = "Temperature deviation (standard deviations)"
)
iplot(
  m_bins_level_signed,
  i.select = 4,
  main = "Signed precipitation-deviation bins",
  xlab = "Precipitation deviation (standard deviations)"
)

etable(
  m_bins_signed,
  m_bins_level_signed,
  headers = c("Signed deviations", "Level and signed bins"),
  fitstat = ~n + r2 + wr2
)
```

### Carleton-style heterogeneity by income and long-run climate

Following the heterogeneous-response construction used by Carleton and
coauthors, each annual temperature-bin indicator is interacted linearly with
the region's sample-average log income and sample-average temperature. Each
precipitation-bin indicator is analogously interacted with sample-average log
income and sample-average precipitation. The modifiers are centered over the
estimation sample, so the uninteracted bin coefficients apply at the sample
means. Region fixed effects absorb the modifiers' main effects.

The figures divide regions into terciles of income and long-run climate, then
evaluate the estimated response at the mean income and climate within each of
the nine cells. All interactions are estimated jointly; the signed anomaly bins
remain in the model as additive deviation controls.

```{r test-carleton-bin-heterogeneity, fig.width=12, fig.height=8}
model_bins_level_signed_heterogeneous <- paste(
  model_bins_level_signed,
  "+ i(TM_bin_level, log_gdp_av_c, ref = \"10-15\")",
  "+ i(TM_bin_level, mean_TM_sample_c, ref = \"10-15\")",
  "+ i(RR_bin_level, log_gdp_av_c, ref = \"0.5-1\")",
  "+ i(RR_bin_level, mean_RR_sample_c, ref = \"0.5-1\")"
)
f_bins_level_signed_heterogeneous <- make_formula(
  model_bins_level_signed_heterogeneous
)
m_bins_level_signed_heterogeneous <- fit_climate_model(
  model_bins_level_signed_heterogeneous
)

temperature_scenarios <- make_carleton_scenarios(
  data,
  climate_var = "mean_TM_sample",
  climate_centered_var = "mean_TM_sample_c",
  climate_group_labels = c("Cold", "Temperate", "Hot")
)
precipitation_scenarios <- make_carleton_scenarios(
  data,
  climate_var = "mean_RR_sample",
  climate_centered_var = "mean_RR_sample_c",
  climate_group_labels = c("Dry", "Moderate", "Wet")
)

temperature_carleton_response <- carleton_bin_response(
  m_bins_level_signed_heterogeneous,
  model_data = data,
  bin_var = "TM_bin_level",
  reference_bin = "10-15",
  income_var = "log_gdp_av_c",
  climate_var = "mean_TM_sample_c",
  scenarios = temperature_scenarios
)
precipitation_carleton_response <- carleton_bin_response(
  m_bins_level_signed_heterogeneous,
  model_data = data,
  bin_var = "RR_bin_level",
  reference_bin = "0.5-1",
  income_var = "log_gdp_av_c",
  climate_var = "mean_RR_sample_c",
  scenarios = precipitation_scenarios
)

plot_carleton_response(
  temperature_carleton_response,
  "Temperature-bin response by average income and temperature",
  "Annual mean temperature (degrees Celsius)"
)
plot_carleton_response(
  precipitation_carleton_response,
  "Precipitation-bin response by average income and precipitation",
  "Annual precipitation (metres)"
)

etable(
  m_bins_level_signed,
  m_bins_level_signed_heterogeneous,
  headers = c("Additive bins", "Income-climate interactions"),
  fitstat = ~n + r2 + wr2
)
```
## GDP growth during extreme temperature and precipitation shocks

The violins compare residual GDP-per-capita growth in years with shocks within
2.5 standard deviations of a region's moving climate baseline against the very
hot/wet and very cold/dry tails. Growth is first residualized using region and
year fixed effects plus region-specific linear year trends, matching the
`GID_1 + year + GID_1[year]` convention. Temperature and precipitation are
shown in separate panels because a region-year can belong to different tail
categories for the two climate variables.

```{r violin-gdp-growth-extreme-shocks, fig.width=10, fig.height=5.5}
shock_levels <- c(
  "Within +/-2.5 SD",
  "> 2.5 SD (very hot/wet)",
  "< -2.5 SD (very cold/dry)"
)

gdp_growth_fe_data <- data %>%
  ungroup() %>%
  filter(
    is.finite(dlgrp_pc_usd),
    !is.na(GID_1),
    !is.na(year)
  )

gdp_growth_fe_model <- fixest::feols(
  dlgrp_pc_usd ~ 1 | GID_1 + year + GID_1[year],
  data = gdp_growth_fe_data
)

gdp_growth_fe_data <- gdp_growth_fe_data %>%
  mutate(
    # Keep an NA placeholder for observations removed internally by fixest
    # (for example, fixed-effect singletons) so rows remain aligned.
    gdp_growth_residual = stats::residuals(
      gdp_growth_fe_model,
      na.rm = FALSE
    )
  )

gdp_growth_by_shock <- dplyr::bind_rows(
  gdp_growth_fe_data %>%
    transmute(
      climate_variable = "Temperature",
      shock = zTM,
      gdp_growth = gdp_growth_residual
    ),
  gdp_growth_fe_data %>%
    transmute(
      climate_variable = "Precipitation",
      shock = zRR,
      gdp_growth = gdp_growth_residual
    )
) %>%
  filter(is.finite(shock), is.finite(gdp_growth)) %>%
  mutate(
    shock_group = case_when(
      shock > 2.5 ~ "> 2.5 SD (very hot/wet)",
      shock < -2.5 ~ "< -2.5 SD (very cold/dry)",
      TRUE ~ "Within +/-2.5 SD"
    ),
    shock_group = factor(shock_group, levels = shock_levels),
    climate_variable = factor(
      climate_variable,
      levels = c("Temperature", "Precipitation")
    )
  )

shock_group_counts <- gdp_growth_by_shock %>%
  count(climate_variable, shock_group, .drop = FALSE)

shock_group_counts

ggplot(
  gdp_growth_by_shock,
  aes(x = shock_group, y = 100 * gdp_growth, fill = shock_group)
) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = "grey30",
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    fill = "white",
    alpha = 0.8,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ climate_variable) +
  scale_fill_manual(
    values = c(
      "Within +/-2.5 SD" = "grey70",
      "> 2.5 SD (very hot/wet)" = "#D55E00",
      "< -2.5 SD (very cold/dry)" = "#0072B2"
    ),
    drop = FALSE
  ) +
  labs(
    title = "Residual GDP growth by standardized climate shock",
    subtitle = paste(
      "Growth residualized by region and year fixed effects",
      "and region-specific linear trends"
    ),
    x = NULL,
    y = "Residual GDP per-capita growth (percentage points)",
    fill = "Shock category"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "none"
  )
```
