---
title: "Tests of Climate Response Functions"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: show
params:
  project_setup_script: null
  climate_velocity_years: 5
  econ_data: "DOSE"
---

# Purpose

This notebook compares alternative functional forms for temperature and
precipitation effects on subnational GDP-per-capita growth. Every model uses the
same climate controls, fixed effects, panel identifiers, and clustered standard
errors so that differences can be attributed to the tested response function.

All functions live in `functions/` and are loaded by `load_functions.R`; the
estimation panel comes from `build_dat()` in `functions/build_dat.R`, which
builds and caches it from the raw sources in `data/`. Set
`project_setup_script` only when project-specific setup is needed first. Set
`econ_data` to `DOSE` (DOSE v2.14) or `KUMMU` (gridded Kummu 2025 aggregated to
GADM 4.1 ADM1) for GADM1, or `PWT` (PWT 11.0) or `WB` (World Bank WDI) for
GADM0; the matching climate resolution is selected automatically.

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "arrow", "curl", "dplyr", "fixest", "ggplot2", "jsonlite", "purrr",
  "readxl", "splines", "tibble", "tidyselect", "zoo"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages)) {
  stop(
    "Install missing packages before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

source("load_functions.R")

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

if (!is.null(params$project_setup_script) &&
    length(params$project_setup_script) == 1L &&
    !is.na(params$project_setup_script) &&
    nzchar(params$project_setup_script)) {
  source(params$project_setup_script, local = globalenv())
}
```

## Shared specification

The common specification contains quadratic long-run temperature and
precipitation terms. All tests add alternative weather-response terms to this
base while retaining the same fixed effects.

```{r shared-specification}
fixed_effects <- "year + GID_1[year]"
pan_id <- c("GID_1", "year")
outcome <- "dlgrp_pc_usd"

base_climate <- "TM + TM:mean_TM_all + RR + RR:mean_RR_all"
model <- paste(base_climate, "+ zRR^2 + zTM^2")

ECON_YEAR_MIN <- 1900

CLIMATE_VELOCITY_YEARS <- as.integer(params$climate_velocity_years)

if (length(CLIMATE_VELOCITY_YEARS) != 1L ||
    is.na(CLIMATE_VELOCITY_YEARS) ||
    CLIMATE_VELOCITY_YEARS < 2L) {
  stop("params$climate_velocity_years must be an integer of at least 2.")
}

# Panel settings; see panel_config() in functions/build_dat.R for the climate
# source, weighting, rolling windows, reference periods and Hamilton filter.
# Enough pre-estimation climate years are retained to calculate the first
# trailing climate-velocity window, as well as the six-year model lags.
panel_cfg <- panel_config(
  econ_year_min = ECON_YEAR_MIN,
  clim_history_years = max(6L, CLIMATE_VELOCITY_YEARS - 1L)
)

# The notebook's specification bound to the shared estimator in
# functions/model_spec.R.
make_formula <- function(terms) {
  make_panel_formula(terms, outcome = outcome, fixed_effects = fixed_effects)
}

fit_climate_model <- function(terms, model_data = data) {
  fit_panel_model(
    terms,
    model_data,
    outcome = outcome,
    fixed_effects = fixed_effects,
    panel_id = pan_id,
    cluster = ~GID_1
  )
}

# Retained as `f` for the split-sample tests and interactive use.
f <- make_formula(model)
```

# Data preparation

The economic sample starts in `r ECON_YEAR_MIN`. Earlier climate observations
are kept as lag carriers and to calculate climate velocity at the beginning of
the estimation period.

```{r build-data}
# build_dat() joins the economic panel to the derived climate moments and adds
# the weather-response variables; both halves are built from raw sources and
# cached under data/cache (see functions/build_dat.R).
data <- build_dat(econ_data = params$econ_data, config = panel_cfg)

data_iso3 <- build_dat(econ_data = "PWT", config = panel_cfg)
```

## Weather-response variables

Weather deviations are standardized by each region's long-run variability.
The derived variables support the piecewise-linear and discrete-bin tests below.

```{r derive-weather-variables}
table(data$TM_bin_signed, useNA = "ifany")
table(data$RR_bin_signed, useNA = "ifany")
table(data$TM_bin_level, useNA = "ifany")
table(data$RR_bin_level, useNA = "ifany")
table(data_iso3$TM_bin_signed, useNA = "ifany")
table(data_iso3$RR_bin_signed, useNA = "ifany")
```

# Test: Correlation between fine temperature and precipitation bins

The Spearman test treats the fine bins as ordered categories. The bin-pair
matrix reports correlations between the indicator for each temperature bin and
the indicator for each precipitation bin.

```{r test-fine-bin-correlation}
fine_bin_data <- data %>%
  filter(
    !is.na(dlgrp_pc_usd),
    !is.na(TM_bin_fine),
    !is.na(RR_bin_fine)
  ) %>%
  transmute(
    TM_bin_fine,
    RR_bin_fine,
    TM_bin_rank = as.integer(TM_bin_fine),
    RR_bin_rank = as.integer(RR_bin_fine)
  )

fine_bin_correlation_test <- cor.test(
  fine_bin_data$TM_bin_rank,
  fine_bin_data$RR_bin_rank,
  method = "spearman",
  exact = FALSE
)

fine_bin_correlation_test

TM_bin_indicators <- model.matrix(
  ~ TM_bin_fine - 1,
  data = fine_bin_data
)
RR_bin_indicators <- model.matrix(
  ~ RR_bin_fine - 1,
  data = fine_bin_data
)

colnames(TM_bin_indicators) <- paste0(
  "TM_",
  levels(fine_bin_data$TM_bin_fine)
)
colnames(RR_bin_indicators) <- paste0(
  "RR_",
  levels(fine_bin_data$RR_bin_fine)
)

fine_bin_pair_correlations <- cor(
  TM_bin_indicators,
  RR_bin_indicators
)

round(fine_bin_pair_correlations, 3)
```


# Test 1: Quadratic weather deviations

This benchmark asks whether squared standardized temperature and precipitation
deviations add information beyond the main BHM climate specification. The two
models use the same estimation sample. A clustered joint Wald test evaluates
the null that both added deviation coefficients are zero.

IDEAS:
- transform the dev variable to z-zmean/(2*zsd) to weight less lower deviation as they show no statistical effects in bins
- main specification should be this + distributed lag models of past SDs?
- check last chatGPT chat for suggestions on spells

```{r test-absolute}

f_bhm <- make_formula(base_climate)
m_bhm <- fit_climate_model(
  base_climate
)

model_base <- paste(
  base_climate,
  "+ zTM^2 + zRR^2"
)

f_base <- make_formula(model_base)
m_base <- fit_climate_model(
  model_base
)

m_base_iso3 <- fit_climate_model(
  model_base,
  model_data = data_iso3
)

deviation_information_test <- fixest::wald(
  m_base,
  keep = "^z(TM|RR^2)$",
  print = FALSE
)

etable(
  m_base_iso3,
  m_bhm,
  m_base,
  headers = c("Iso3", "Naive", "Full"),
  fitstat = ~n + r2 + wr2 + aic + bic
)

```



```{r test-quadratic}
model_abs <- paste(
  base_climate,
  "+ abs(zTM) + abs(zRR)"
)
m_abs <- fit_climate_model(model_abs)

etable(m_base,m_abs, fitstat = ~n + r2 + wr2)

```


# Test 1: lags

How much memory is in the system?

```{r test lags}

data <- data |> group_by(GID_1) |> 
  mutate(dTM=TM-lag_by_year(TM,year),dRR=RR-lag_by_year(RR,year)) |>
  ungroup()

model_lags <- paste(
#  "l(TM,0:5) + l(TM,0:5):mean_TM_all",
#  "+ l(RR,0:5) + l(RR,0:5):mean_RR_all",
  "+ l(zRR^2,0:10) + l(zTM^2,0:10)"
)

f_lags <- make_formula(model_lags)
m_lags <- fit_climate_model(model_lags)

etable(m_lags, fitstat = ~n + r2 + wr2)
  

```

```{r plot-distributed-lags, fig.width=10, fig.height=6}
# Each lag coefficient and the cumulative response through that lag; see
# distributed_lag_effects() in functions/response_functions.R.
lag_confidence_level <- 0.95
distributed_lag_data <- distributed_lag_effects(
  m_lags,
  conf_level = lag_confidence_level
)
plot_distributed_lags(distributed_lag_data, conf_level = lag_confidence_level)
```

```{r test windows}

model_windows <- paste(
  base_climate,
  " + abs((TM - mean_TM_lag_30) / sd_TM_lag_30)   + abs((TM - mean_TM_lag_5) / sd_TM_lag_5)",
  " + abs((RR - mean_RR_lag_30) / sd_RR_lag_30)   + abs((RR - mean_RR_lag_5) / sd_RR_lag_5)"
)

f_windows <- make_formula(model_windows)
m_windows <- fit_climate_model(model_windows)

etable(m_windows, fitstat = ~n + r2 + wr2)
  

```


# Test 3: Coarse unsigned anomaly bins

This nonparametric specification groups absolute anomalies into three broad
ranges, allowing moderate and extreme weather to have distinct effects.

```{r test-coarse-bins}
model_bin <- paste(base_climate, "+ TM_bin + RR_bin")
f_bin <- make_formula(model_bin)
m_bin <- fit_climate_model(model_bin)

etable(m_bin, fitstat = ~n + r2 + wr2)
```

# Test 4: Fine unsigned anomaly bins

This test increases resolution to half-standard-deviation intervals and uses
the 0--0.5 standard-deviation category as the reference for both variables.

```{r test-fine-bins}
model_bins <- paste(
  base_climate,
  "+ i(TM_bin_fine, ref = \"0-.5\")",
  "+ i(RR_bin_fine, ref = \"0-.5\")"
)
f_bins <- make_formula(model_bins)
m_bins <- fit_climate_model(model_bins)

etable(m_bins, fitstat = ~n + r2 + wr2)
```


```{r moving averages interaction}

model_moving <- paste(
  "TM + TM:mean_TM_lag_30 + RR + RR:mean_RR_lag_30",
  "+ zRR^2 + zTM^2"
)

m_move <- fit_climate_model(model_moving)

etable(m_bhm,m_base,m_move)


```

# Test 5: Variable-specific fine bins

Estimating the fine-bin response for one weather variable at a time makes its
shape easier to inspect without simultaneously estimating the other bin set.

```{r test-variable-specific-fine-bins, fig.width=8, fig.height=5}
model_bins_fine_TM <- paste(
  base_climate,
  "+ i(TM_bin_fine, ref = \"0-.5\")"
)
f_bins_fine_TM <- make_formula(model_bins_fine_TM)
m_bins_fine_TM <- fit_climate_model(model_bins_fine_TM)

model_bins_fine_RR <- paste(
  base_climate,
  "+ i(RR_bin_fine, ref = \"0-.5\")"
)
f_bins_fine_RR <- make_formula(model_bins_fine_RR)
m_bins_fine_RR <- fit_climate_model(model_bins_fine_RR)

iplot(m_bins_fine_TM, main = "Temperature anomaly bins")
iplot(m_bins_fine_RR, main = "Precipitation anomaly bins")
```

# Test 6: Signed anomaly bins

Unsigned bins impose the same response on positive and negative deviations.
This test instead distinguishes unusually cold from hot years and unusually dry
from wet years, with deviations between -0.5 and 0.5 standard deviations as the
reference category.

```{r test-signed-bins, fig.width=8, fig.height=5}

model_bins_signed <- paste(
  base_climate,
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
)
f_bins_signed <- make_formula(model_bins_signed)
m_bins_signed <- fit_climate_model(model_bins_signed)

iplot(m_bins_signed, main = "Signed temperature anomaly bins")
iplot(m_bins_signed, i.select=2,main = "Signed precipitation anomaly bins")

```

## Temperature and precipitation level bins with signed deviations

This specification adds nonparametric bins for annual temperature and
precipitation levels to the joint signed-deviation model. Temperature bins use
5-degree Celsius intervals and precipitation bins use 0.5-metre intervals. The
most populated central intervals (10--15 degrees Celsius and 0.5--1 metre) are
the level-bin references; the signed-deviation reference remains -0.5 to 0.5
within-region standard deviations. Consequently, every plotted coefficient is
conditional on the common BHM climate controls and on the other three bin sets.

```{r test-level-and-signed-bins, fig.width=8, fig.height=5, fig.show='hold', out.width='50%'}
model_bins_level_signed <- paste(
  "+ i(TM_bin_level)",
  "+ i(RR_bin_level)",
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
)
f_bins_level_signed <- make_formula(model_bins_level_signed)
m_bins_level_signed <- fit_climate_model(model_bins_level_signed)

iplot(
  m_bins_level_signed,
  i.select = 1,
  main = "Annual temperature bins",
  xlab = "Annual mean temperature (degrees Celsius)"
)
iplot(
  m_bins_level_signed,
  i.select = 2,
  main = "Annual precipitation bins",
  xlab = "Annual precipitation (metres)"
)
iplot(
  m_bins_level_signed,
  i.select = 3,
  main = "Signed temperature-deviation bins",
  xlab = "Temperature deviation (standard deviations)"
)
iplot(
  m_bins_level_signed,
  i.select = 4,
  main = "Signed precipitation-deviation bins",
  xlab = "Precipitation deviation (standard deviations)"
)

etable(
  m_bins_signed,
  m_bins_level_signed,
  headers = c("Signed deviations", "Level and signed bins"),
  fitstat = ~n + r2 + wr2
)
```

### Carleton-style heterogeneity by income and long-run climate

Following the heterogeneous-response construction used by Carleton and
coauthors, each annual temperature-bin indicator is interacted linearly with
the region's sample-average log income and sample-average temperature. Each
precipitation-bin indicator is analogously interacted with sample-average log
income and sample-average precipitation. The modifiers are centered over the
estimation sample, so the uninteracted bin coefficients apply at the sample
means. Region fixed effects absorb the modifiers' main effects.

The figures divide regions into terciles of income and long-run climate, then
evaluate the estimated response at the mean income and climate within each of
the nine cells. All interactions are estimated jointly; the signed anomaly bins
remain in the model as additive deviation controls.

```{r test-carleton-bin-heterogeneity, fig.width=12, fig.height=8}
model_bins_level_signed_heterogeneous <- paste(
  model_bins_level_signed,
  "+ i(TM_bin_level, log_gdp_av_c, ref = \"10-15\")",
  "+ i(TM_bin_level, mean_TM_sample_c, ref = \"10-15\")",
  "+ i(RR_bin_level, log_gdp_av_c, ref = \"0.5-1\")",
  "+ i(RR_bin_level, mean_RR_sample_c, ref = \"0.5-1\")"
)
f_bins_level_signed_heterogeneous <- make_formula(
  model_bins_level_signed_heterogeneous
)
m_bins_level_signed_heterogeneous <- fit_climate_model(
  model_bins_level_signed_heterogeneous
)

temperature_scenarios <- make_carleton_scenarios(
  data,
  climate_var = "mean_TM_sample",
  climate_centered_var = "mean_TM_sample_c",
  climate_group_labels = c("Cold", "Temperate", "Hot")
)
precipitation_scenarios <- make_carleton_scenarios(
  data,
  climate_var = "mean_RR_sample",
  climate_centered_var = "mean_RR_sample_c",
  climate_group_labels = c("Dry", "Moderate", "Wet")
)

temperature_carleton_response <- carleton_bin_response(
  m_bins_level_signed_heterogeneous,
  model_data = data,
  bin_var = "TM_bin_level",
  reference_bin = "10-15",
  income_var = "log_gdp_av_c",
  climate_var = "mean_TM_sample_c",
  scenarios = temperature_scenarios
)
precipitation_carleton_response <- carleton_bin_response(
  m_bins_level_signed_heterogeneous,
  model_data = data,
  bin_var = "RR_bin_level",
  reference_bin = "0.5-1",
  income_var = "log_gdp_av_c",
  climate_var = "mean_RR_sample_c",
  scenarios = precipitation_scenarios
)

plot_carleton_response(
  temperature_carleton_response,
  "Temperature-bin response by average income and temperature",
  "Annual mean temperature (degrees Celsius)"
)
plot_carleton_response(
  precipitation_carleton_response,
  "Precipitation-bin response by average income and precipitation",
  "Annual precipitation (metres)"
)

etable(
  m_bins_level_signed,
  m_bins_level_signed_heterogeneous,
  headers = c("Additive bins", "Income-climate interactions"),
  fitstat = ~n + r2 + wr2
)
```

# PREFERRED SPECIFICATION:

```{r test-double deviation}

data <- data |> 
    mutate(zzTM = pmax(1.5,abs((TM - mean_TM_lag_30) / (sd_TM_lag_30)))-1.5 ) 

m_pref <- fit_climate_model(paste(base_climate,"+ zzTM^2 + abs_zRRm"))

etable(m_base, m_pref, fitstat = ~n + r2 + wr2)

```


```{r signed, national}

m_bins_signed_iso3 <- fit_climate_model(model_bins_signed,model_data=data_iso3)
m_quad_iso3 <- fit_climate_model(model_base,model_data=data_iso3)


iplot(m_bins_signed, m_bins_signed_iso3, main = "Signed temperature anomaly bins")

```


# signed, divided by mean temperature and precip
```{r temperature and precip robust}
data_income <- data %>%
  group_by(GID_0) %>%
  mutate(av_y = mean(grp_pc_usd, na.rm = TRUE)) %>%
  ungroup()

income_cutoffs <- quantile(
  data_income$av_y,
  c(0.25, 0.5, 0.75),
  na.rm = TRUE
)
mean_TM_cutoff <- median(data_income$mean_TM_all, na.rm = TRUE)
mean_RR_cutoff <- median(data_income$mean_RR_all, na.rm = TRUE)

data_signed_groups <- data_income %>%
  mutate(
    mean_temperature_group = factor(
      if_else(mean_TM_all < mean_TM_cutoff, "Cold", "Hot"),
      levels = c("Cold", "Hot")
    ),
    mean_precipitation_group = factor(
      if_else(mean_RR_all < mean_RR_cutoff, "Dry", "Wet"),
      levels = c("Dry", "Wet")
    ),
    income_group = factor(
      if_else(av_y <= income_cutoffs[[1]], "Poor", "Rich"),
      levels = c("Poor", "Rich")
    )
  )

m_signed_tlo <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_temperature_group == "Cold"),
  panel.id = pan_id,
  vcov = ~GID_1
)

m_signed_thi <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_temperature_group == "Hot"),
  panel.id = pan_id,
  vcov = ~GID_1
)


f_bins_signed_int <- make_formula(paste(
    base_climate,
  "+ i(TM_bin_signed, mean_temperature_group, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
))


m_signed_plo <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_precipitation_group == "Dry"),
  panel.id = pan_id,
  vcov = ~GID_1
)


m_signed_phi <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_precipitation_group == "Wet"),
  panel.id = pan_id,
  vcov = ~GID_1
)


m_signed_ylo <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(income_group == "Poor"),
  panel.id = pan_id,
  vcov = ~GID_1
)


m_signed_ymhi <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(income_group == "Rich"),
  panel.id = pan_id,
  vcov = ~GID_1
)

make_common_base_group_formula <- function(group_var) {
  as.formula(
    paste(
      outcome,
      "~",
      base_climate,
      paste0("+ i(TM_bin_signed, ", group_var, ", ref = 0)"),
      paste0("+ i(RR_bin_signed, ", group_var, ", ref = 0)"),
      "|",
      fixed_effects
    )
  )
}

m_signed_t_common_base <- fixest::feols(
  make_common_base_group_formula("mean_temperature_group"),
  data = data_signed_groups,
  panel.id = pan_id,
  vcov = ~GID_1
)

m_signed_p_common_base <- fixest::feols(
  make_common_base_group_formula("mean_precipitation_group"),
  data = data_signed_groups,
  panel.id = pan_id,
  vcov = ~GID_1
)

m_signed_income_common_base <- fixest::feols(
  make_common_base_group_formula("income_group"),
  data = data_signed_groups,
  panel.id = pan_id,
  vcov = ~GID_1
)

etable(
  m_bins_signed,
  m_signed_tlo,
  m_signed_thi,
  m_signed_plo,
  m_signed_phi,
  m_signed_ylo,
  m_signed_ymhi,
  headers = c("All", "Cold", "Hot", "Dry", "Wet","Poor","Rich"),
  fitstat = ~n + r2 + wr2
)

etable(
  m_signed_t_common_base,
  m_signed_p_common_base,
  m_signed_income_common_base,
  headers = c("Mean temperature", "Mean precipitation", "Income"),
  fitstat = ~n + r2 + wr2
)

mean_temperature_bin_data <- interacted_bin_coefs(
  m_signed_t_common_base,
  bin_var = "TM_bin_signed",
  group_var = "mean_temperature_group",
  group_levels = levels(data_signed_groups$mean_temperature_group)
) %>%
  mutate(
    bin_label = numeric_bin_factor(bin_label),
    group = factor(group, levels = c("Cold", "Hot"))
  )

signed_temperature_bins_by_mean_temperature <- plot_group_bins(
  mean_temperature_bin_data,
  colors = c(Cold = "#0072B2", Hot = "#D55E00"),
  shapes = c(Cold = 16, Hot = 17),
  title = "Signed temperature-bin coefficients by mean temperature",
  group_label = "Country group",
  x_label = "Signed temperature anomaly bin (standard deviations)"
)

signed_temperature_bins_by_mean_temperature

iplot(
  m_signed_thi,
  m_signed_tlo,
  i.select = 1,
  main = "Signed temperature bins by mean temperature"
)
iplot(
  m_signed_phi,
  m_signed_plo,
  i.select = 2,
  main = "Signed precipitation bins by mean precipitation"
)


iplot(
  m_signed_ymhi,
  m_signed_ylo,
  i.select = 1,
  main = "Signed temperature bins by income"
)


iplot(
  m_signed_ymhi,
  m_signed_ylo,
  i.select = 2,
  main = "Signed precipitation bins by income"
)

```

# Test: Signed temperature and precipitation bins by climate zone

`mean_kg` takes the integer values 1--5 and is constant over time within every
region with nonmissing data in the source file. Notable entries reveal the five broad
Koppen--Geiger climate families: Indonesia and Singapore are coded 1; Saudi
Arabia, Egypt, and the United Arab Emirates are coded 2; France and the United
Kingdom are coded 3; Mongolia and most Russian regions are coded 4; and
Greenland is coded 5. The resulting interpretation is tropical, arid,
temperate, continental, and polar, respectively. Because the polar class has
too few estimation observations, source code 5 is merged into code 4 below and
the combined regression group is labeled Continental/Polar.

```{r load-koppen-geiger-zones}
kg_path <- file.path("econometrics", "data", "kg_data.csv")
if (!file.exists(kg_path)) {
  kg_path <- file.path("data", "kg_data.csv")
}
if (!file.exists(kg_path)) {
  stop("Cannot find econometrics/data/kg_data.csv.")
}

kg_zone_key <- data.frame(
  mean_kg = 1:4,
  climate_zone = c(
    "1 Tropical (A)",
    "2 Arid (B)",
    "3 Temperate (C)",
    "4 Continental/Polar (D/E)"
  )
)

kg_data <- read.csv(
  kg_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
) |>  select(-1) %>%
  transmute(
    GID_0 = as.character(GID_0),
    GID_1 = as.character(GID_1),
    year = as.integer(year),
    mean_kg_source = as.integer(mean_kg),
    mean_kg = if_else(mean_kg_source == 5L, 4L, mean_kg_source)
  ) %>%
  left_join(kg_zone_key, by = "mean_kg") %>%
  mutate(
    climate_zone = factor(
      climate_zone,
      levels = kg_zone_key$climate_zone
    )
  )

stopifnot(
  !anyDuplicated(kg_data[c("GID_0", "GID_1", "year")]),
  all(na.omit(unique(kg_data$mean_kg_source)) %in% 1:5),
  all(na.omit(unique(kg_data$mean_kg)) %in% kg_zone_key$mean_kg)
)

kg_zone_key

data_kg <- data %>%
  left_join(
    kg_data %>%
      select(GID_0, GID_1, year, mean_kg_source, mean_kg, climate_zone),
    by = c("GID_0", "GID_1", "year")
  )

kg_match_summary <- data_kg %>%
  filter(!is.na(.data[[outcome]])) %>%
  summarise(
    observations = n(),
    matched = sum(!is.na(climate_zone)),
    match_share = matched / observations
  )

kg_zone_sample <- data_kg %>%
  filter(!is.na(.data[[outcome]]), !is.na(climate_zone)) %>%
  group_by(climate_zone) %>%
  summarise(
    observations = n(),
    regions = n_distinct(GID_1),
    countries = n_distinct(GID_0),
    .groups = "drop"
  )

kg_match_summary
kg_zone_sample
```

Each climate-zone regression jointly estimates the signed temperature and
precipitation bins using the same controls, fixed effects, and region-clustered
standard errors as the pooled signed-bin model.

```{r test-signed-bins-by-climate-zone, fig.width=10, fig.height=6}
f_bins_signed_zone <- make_formula(paste(
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
))

m_bins_signed_by_climate <- fixest::feols(
  f_bins_signed_zone,
  data = data_kg %>% filter(!is.na(climate_zone)),
  panel.id = pan_id,
  vcov = ~GID_1,
  split = ~climate_zone
)

m_bins_signed_by_climate_with_i <- feols(
  dlgrp_pc_usd ~
    i(TM_bin_signed, climate_zone, ref = 0) +
    i(RR_bin_signed, climate_zone, ref = 0)
  | year +
    GID_1 +
    GID_1[year],
  data = data_kg %>% filter(!is.na(climate_zone)),
  panel.id = pan_id,
  vcov = ~GID_1
)

etable(
  m_bins_signed_by_climate,
  m_bins_signed_by_climate_with_i,
  fitstat = ~n + r2 + wr2
)

climate_zone_colors <- c(
  "#009E73", # Tropical
  "#E69F00", # Arid
  "#0072B2", # Temperate
  "#CC79A7"  # Continental/Polar
)
climate_zone_pch <- c(20, 17, 15, 18)
climate_zone_legend <- sub(
  "^[0-9]+ ",
  "",
  levels(data_kg$climate_zone)
)
names(climate_zone_colors) <- climate_zone_legend
names(climate_zone_pch) <- climate_zone_legend

# Split-sample bins, labelled by climate zone without the numeric code prefix.
plot_climate_zone_bins <- function(models, i_select, title) {
  plot_data <- split_bin_coefs(models, i_select) %>%
    mutate(
      bin_label = numeric_bin_factor(bin_label),
      group = factor(sub("^[0-9]+ ", "", group), levels = climate_zone_legend)
    )

  plot_group_bins(
    plot_data,
    colors = climate_zone_colors,
    shapes = climate_zone_pch,
    title = title,
    group_label = "Climate zone",
    facet = TRUE,
    ylim = c(-0.1, 0.05)
  )
}

plot_climate_zone_bins(
  m_bins_signed_by_climate,
  i_select = 1,
  title = paste(
    "Signed temperature bins by climate zone",
    "(shared climate coefficients)"
  )
)

plot_climate_zone_bins(
  m_bins_signed_by_climate,
  i_select = 2,
  title = paste(
    "Signed precipitation bins by climate zone",
    "(shared climate coefficients)"
  )
)
```

# Test 7: Heterogeneity by climate variability

These split-sample regressions test whether the baseline deviation response
differs between regions below and above the median long-run variability of
precipitation or temperature.

```{r test-climate-variability-splits}
mplow <- fixest::feols(
  f,
  data = data %>% filter(sd_RR_all < median(sd_RR_all, na.rm = TRUE)),
  panel.id = pan_id
)
mphi <- fixest::feols(
  f,
  data = data %>% filter(sd_RR_all > median(sd_RR_all, na.rm = TRUE)),
  panel.id = pan_id
)

mtlow <- fixest::feols(
  f,
  data = data %>% filter(sd_TM_all < median(sd_TM_all, na.rm = TRUE)),
  panel.id = pan_id
)
mthi <- fixest::feols(
  f,
  data = data %>% filter(sd_TM_all > median(sd_TM_all, na.rm = TRUE)),
  panel.id = pan_id
)

etable(
  mplow,
  mphi,
  mtlow,
  mthi,
  headers = c("Low RR SD", "High RR SD", "Low TM SD", "High TM SD"),
  fitstat = ~n + r2 + wr2
)
```

# Test 8: Heterogeneity by average income

The final test compares the baseline response in the bottom and top quartiles of
region-level average GDP per capita.

```{r test-income-splits}
mylow <- fixest::feols(
  f,
  data = data_income %>% filter(av_y < income_cutoffs[[2]]),
  panel.id = pan_id
)
myhi <- fixest::feols(
  f,
  data = data_income %>% filter(av_y >= income_cutoffs[[2]]),
  panel.id = pan_id
)

etable(
  mylow,
  myhi,
  headers = c("Bottom income quartile", "Top income quartile"),
  fitstat = ~n + r2 + wr2
)
```

# Test 9: Asymmetric temperature--precipitation interactions

This specification allows positive and negative temperature and precipitation
deviations to have separate effects. The four interactions additionally allow
the response to differ across the hot--wet, hot--dry, cold--wet, and cold--dry
quadrants. Each component is expressed as a non-negative anomaly magnitude.

```{r test-asymmetric-interactions}


model_asymmetric_interactions <- paste(
  base_climate,
  "+ abs_zTMp^2 + abs_zTMm^2 + abs_zRRp + abs_zRRm",
  "+ abs_zTMp^2:abs_zRRp + abs_zTMp^2:abs_zRRm",
  "+ abs_zTMm^2:abs_zRRp + abs_zTMm^2:abs_zRRm"
)

f_asymmetric_interactions <- make_formula(model_asymmetric_interactions)
m_asymmetric_interactions <- fit_climate_model(model_asymmetric_interactions)

m_asym_low <- fixest::feols(
  f_asymmetric_interactions,
  data = data_income %>% filter(av_y < income_cutoffs[[2]]),
  panel.id = pan_id,
    cluster = ~GID_1
)
m_asym_hi <- fixest::feols(
  f_asymmetric_interactions,
  data = data_income %>% filter(av_y >= income_cutoffs[[2]]),
  panel.id = pan_id, 
    cluster = ~GID_1
)
etable(
  m_asymmetric_interactions,
  m_asym_low,
  m_asym_hi,
  fitstat = ~n + r2 + wr2
)
```

# Test 10: Climate velocity during warming and cooling runs

Here, `climate_velocity` is a short-run temporal warming rate rather than a
spatial velocity. For year $t$, it is estimated from the fixed trailing
`r params$climate_velocity_years`-year window ending in $t$. The window therefore
uses no future observations. The slope is expressed in degrees Celsius per
decade and is calculated independently of whether year $t$ belongs to a warming
run.

A warming run is a maximal sequence of at least two consecutive calendar years
with strictly increasing annual temperature. Position 1 is the year immediately
before the first positive change, and position 2 is the first warmer year.
`warming_run_rate` records the separate slope
estimated only from a run's observations. `warming_run_length` records the
year's position within its run, while `warming_run_total_length` preserves the
run's eventual complete length. Single-year events are coded as 1 rather than
0, making them equivalent to the first year of a subsequent run. Cooling runs
and their corresponding fields are defined symmetrically from strictly
decreasing annual temperature. A local peak or trough can therefore be the
final year of one run and the first year of the opposite run.
`cumulative_warming` and `cumulative_cooling` sum the annual temperature
changes belonging to the corresponding run through the current year. They are
zero in position 1; cumulative warming is positive and cumulative cooling is
negative thereafter.

The fixed-baseline deviations hold the lagged 20-year temperature mean and
standard deviation at their position-1 values throughout each run. Because a
turning point belongs to both an ending run and a new run, the standardized
deviations are retained separately as `zTM_warming_f` and `zTM_cooling_f`.
Their one-sided magnitudes are `zTMp_f` and `zTMm_f`. Outside the corresponding
run, and at position 1, these equal the original moving-baseline deviations.

```{r construct-climate-velocity}
# Run descriptors and trailing climate velocity; see add_climate_runs() in
# functions/climate_runs.R. check_climate_runs() asserts that every run includes
# its first year and at least two observations, and that the fixed-baseline
# deviations and cumulative changes are as defined above at position 1. It
# warns about flat-temperature years that belong to neither run.
climate_velocity_hist <- add_climate_runs(
  data,
  velocity_years = CLIMATE_VELOCITY_YEARS,
  baseline_mean = "mean_TM_lag_30",
  baseline_sd = "sd_TM_lag_30"
)
check_climate_runs(climate_velocity_hist)

warming_run_summary <- climate_velocity_hist %>%
  filter(warming_run) %>%
  group_by(GID_1, warming_run_id) %>%
  summarise(
    warming_run_total_length = first(warming_run_total_length),
    warming_run_rate = first(warming_run_rate),
    .groups = "drop"
  ) %>%
  arrange(GID_1, warming_run_id)

warming_run_summary %>%
  count(warming_run_total_length, name = "runs")

cooling_run_summary <- climate_velocity_hist %>%
  filter(cooling_run) %>%
  group_by(GID_1, cooling_run_id) %>%
  summarise(
    cooling_run_total_length = first(cooling_run_total_length),
    cooling_run_rate = first(cooling_run_rate),
    .groups = "drop"
  ) %>%
  arrange(GID_1, cooling_run_id)

cooling_run_summary %>%
  count(cooling_run_total_length, name = "runs")

climate_velocity_hist %>%
  filter(warming_run, !is.na(climate_velocity)) %>%
  ggplot(aes(
    x = climate_velocity_warming,
    color = factor(warming_run_total_length)
  )) +
  geom_density() +
  labs(color = "Warming-run length")

climate_velocity_hist %>%
  filter(cooling_run, !is.na(climate_velocity)) %>%
  ggplot(aes(
    x = climate_velocity_cooling,
    color = factor(cooling_run_total_length)
  )) +
  geom_density() +
  labs(color = "Cooling-run length")
 
climate_velocity_hist |> filter(warming_run_total_length>=7) |> group_by(GID_1) |> filter(year==min(year)) |>  ungroup() |> select(GID_0,year) |> unique()


climate_velocity_hist |> filter(warming_run_total_length>=6) |> group_by(GID_1) |> filter(year==max(year)) |>  ungroup() |> select(GID_1,year,cumulative_warming) |> unique() |> 
  ggplot() + geom_density(aes(x=cumulative_warming))
```

## regressions for climate velocity

```{r test-climate-velocity}

model_velocity_bool <- paste(
  base_climate, 
  "+ zRR^2 +",
  "+ I(zTMp_f^2) : i(warming_run_length)",
  "+ I(zTMm_f^2) : i(cooling_run_length)",
  "+ l(zTMp_f^2,1:3) + l(zTMm_f^2,1:3)"
)

m_bool <- fit_climate_model(
  model_velocity_bool,
  model_data = climate_velocity_hist
)



model_cumulative_change <- paste(
  base_climate,
  "+ zRR^2 +",
  "+ I(zTMp_f^2) : cumulative_warming",
  "+ I(zTMm_f^2) : cumulative_cooling",
  "+ I(zTMp_f^2) + I(zTMm_f^2)",
  "+ l(zTMp_f,1:3) + l(zTMm_f,1:3)"
)

m_cumulative_change <- fit_climate_model(
  model_cumulative_change,
  model_data = climate_velocity_hist
)

climate_velocity_hist <- climate_velocity_hist %>%
  mutate(
    climate_velocity_warming_bin = bin_nonzero_magnitude(
      climate_velocity_warming
    ),
    climate_velocity_cooling_bin = bin_nonzero_magnitude(
      abs(climate_velocity_cooling)
    ),
    cumulative_warming_bin = bin_nonzero_magnitude(
      cumulative_warming
    ),
    cumulative_cooling_bin = bin_nonzero_magnitude(
      abs(cumulative_cooling)
    )
  )



model_velocity_bin <- paste(
  base_climate,
  paste0(" + i(climate_velocity_warming_bin, I(zTMp_f^2))",
  "+ l(zTMp_f^2,1:3) + l(zTMm_f^2,0:3)"
  )
)

f_velocity <- make_formula(model_velocity_bin)
m_velocity <- fit_climate_model(
  model_velocity_bin,
  model_data = climate_velocity_hist
)

model_cumulative_bin <- paste(
  base_climate,
  paste0(
    "+  ",
    "i(cumulative_warming_bin, I(zTMp_f^2))"
  ),
  paste0(
    "+  ",
    "i(cumulative_cooling_bin, I(zTMm_f^2))",
  "+ l(zTMp_f^2,1:3) + l(zTMm_f^2,1:3)"
  )
)

f_cumulative_bin <- make_formula(model_cumulative_bin)
m_cumulative_bin <- fit_climate_model(
  model_cumulative_bin,
  model_data = climate_velocity_hist
)

etable(
  m_velocity,
  m_cumulative_bin,
  m_bool,
  m_cumulative_change,
  fitstat = ~n + r2 + wr2
)
iplot(
  m_velocity, 
  main = "Temperature response and trailing climate velocity"
)

iplot(
  m_cumulative_bin, i.select=1,
  main = "Temperature response and cumulative run change (warming)"
)

iplot(
  m_cumulative_bin, i.select=2,
  main = "Temperature response and cumulative run change (cooling)"
)

iplot(
  m_bool, 
  main = "Temperature response and consecutive warming years"
)

etable(m_bool, fitstat = ~n + r2 + wr2)

```

## Association between cumulative-change and climate-velocity bins

The association is calculated separately for warming and cooling. Spearman's
rho treats the bins as ordered, while Cramer's V treats them as categorical.
The jointly non-zero results isolate active cumulative change and climate
velocity from their shared zero reference category.

```{r cumulative-velocity-bin-association, fig.width=10, fig.height=5}
bin_association_summary <- dplyr::bind_rows(
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_warming_bin",
    "climate_velocity_warming_bin",
    "Warming",
    "All complete"
  ),
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_warming_bin",
    "climate_velocity_warming_bin",
    "Warming",
    "Both non-zero"
  ),
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_cooling_bin",
    "climate_velocity_cooling_bin",
    "Cooling",
    "All complete"
  ),
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_cooling_bin",
    "climate_velocity_cooling_bin",
    "Cooling",
    "Both non-zero"
  )
)

bin_association_summary

joint_active_bins <- dplyr::bind_rows(
  climate_velocity_hist %>%
    transmute(
      direction = "Warming",
      cumulative_bin = cumulative_warming_bin,
      velocity_bin = climate_velocity_warming_bin
    ),
  climate_velocity_hist %>%
    transmute(
      direction = "Cooling",
      cumulative_bin = cumulative_cooling_bin,
      velocity_bin = climate_velocity_cooling_bin
    )
) %>%
  filter(
    !is.na(cumulative_bin),
    !is.na(velocity_bin),
    as.character(cumulative_bin) != "0",
    as.character(velocity_bin) != "0"
  ) %>%
  count(direction, cumulative_bin, velocity_bin, name = "observations") %>%
  group_by(direction) %>%
  mutate(share = observations / sum(observations)) %>%
  ungroup()

ggplot(
  joint_active_bins,
  aes(x = velocity_bin, y = cumulative_bin, fill = share)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", 100 * share)), size = 3) +
  facet_wrap(~direction, scales = "free") +
  labs(
    x = "Climate-velocity magnitude bin",
    y = "Cumulative-change magnitude bin",
    fill = "Share",
    title = "Joint distribution among non-zero run indicators"
  ) +
  theme_minimal()
```

## test only lags that belong to the same warming run

```{r}


climate_velocity_hist <- climate_velocity_hist %>%
  arrange(GID_1, year) %>%
  mutate(
    zTMp2 = zTMp_f^2,
    zTMm2 = zTMm_f^2
  )

# Generate every feasible lag rather than imposing an arbitrary three-year cap.
max_warming_same_run_lag <- max(
  climate_velocity_hist$warming_run_length,
  na.rm = TRUE
) - 1L
max_cooling_same_run_lag <- max(
  climate_velocity_hist$cooling_run_length,
  na.rm = TRUE
) - 1L

climate_velocity_hist <- add_same_run_lags(
  climate_velocity_hist,
  value_column = "zTMp2",
  run_length_column = "warming_run_length",
  prefix = "zTMp2_l",
  max_lag = max_warming_same_run_lag
) %>%
  add_same_run_lags(
    value_column = "zTMm2",
    run_length_column = "cooling_run_length",
    prefix = "zTMm2_l",
    max_lag = max_cooling_same_run_lag
  )

warming_same_run_lag_terms <- paste0(
  "zTMp2_l",
  seq_len(max_warming_same_run_lag),
  "_same"
)
cooling_same_run_lag_terms <- paste0(
  "zTMm2_l",
  seq_len(max_cooling_same_run_lag),
  "_same"
)
same_run_lag_terms <- c(
  warming_same_run_lag_terms,
  cooling_same_run_lag_terms
)
same_run_lag_formula <- if (length(same_run_lag_terms)) {
  paste("+", paste(same_run_lag_terms, collapse = " + "))
} else {
  ""
}

# A lag is zero whenever looking back that far would leave the current run.
for (lag_number in seq_len(max_warming_same_run_lag)) {
  lag_column <- warming_same_run_lag_terms[[lag_number]]
  stopifnot(all(
    climate_velocity_hist[[lag_column]][
      climate_velocity_hist$warming_run_length <= lag_number
    ] == 0
  ))
}
for (lag_number in seq_len(max_cooling_same_run_lag)) {
  lag_column <- cooling_same_run_lag_terms[[lag_number]]
  stopifnot(all(
    climate_velocity_hist[[lag_column]][
      climate_velocity_hist$cooling_run_length <= lag_number
    ] == 0
  ))
}

model_velocity_bool <- paste(
  base_climate, 
  "+ abs_zRRm + abs_zRRp +",
  "+ I(zTMp_f^2) + I(zTMm_f^2)",
  same_run_lag_formula
)

```


# Test 11: the SD model vs the no SD model


```{r test SD vs no SD}

# Without region-specific scaling: raw deviations from the same lagged 30-year
# mean, divided by one common SD (the estimation-sample average) so every region
# gets bins of the same physical width. Bin codes stay in "average SD" units,
# which keeps the two models' bins directly comparable.
sd_TM_common <- mean(data$sd_TM_lag_30[est_row(data)], na.rm = TRUE)
sd_RR_common <- mean(data$sd_RR_lag_30[est_row(data)], na.rm = TRUE)
data_nosd <- data |>
  mutate(
    TM_bin_signed = signed_anomaly_bin((TM - mean_TM_lag_30) / sd_TM_common),
    RR_bin_signed = signed_anomaly_bin((RR - mean_RR_lag_30) / sd_RR_common)
  )
model_nosd <- paste(
  base_climate,
  paste0(
    "+ i(TM_bin_signed, ref = 0) ",
    "+ i(RR_bin_signed, ref = 0) "
  )
)

m_nosd <- fit_climate_model(
  model_nosd,
  model_data = data_nosd
)

etable(m_bins_signed, m_nosd, fitstat = ~n + r2 + wr2)
iplot(
  m_nosd, m_bins_signed, i.select=1,
  main = "Temperature response"
)
iplot(
  m_nosd, m_bins_signed, i.select=2,
  main = "Precipitation response"
)

```

```{r test adaptation}

ADAPTATION_SPLINE_DF <- 3L

data_adaptation <- data %>%
  group_by(gadm_level, GID_1) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    lag_lgrp_pc_usd_1 = lag_by_year(lgrp_pc_usd, year)
  ) %>%
  ungroup()

# Natural-spline basis of lagged income, stored as scalar columns so fixest can
# interact each one with the i() bins; see add_spline_basis().
income_spline <- add_spline_basis(
  data_adaptation,
  "lag_lgrp_pc_usd_1",
  df = ADAPTATION_SPLINE_DF,
  prefix = "lag_lgrp_spline_"
)
data_adaptation <- income_spline$data
income_spline_fit <- income_spline$basis
spline_columns <- income_spline$columns
valid_income_lag <- is.finite(data_adaptation$lag_lgrp_pc_usd_1)

temperature_adaptation_terms <- paste0(
  "i(TM_bin_signed, ", spline_columns, ", ref = 0)"
)
precipitation_adaptation_terms <- paste0(
  "i(RR_bin_signed, ", spline_columns, ", ref = 0)"
)

model_ada <- paste(
  base_climate,
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)",
  "+", paste(
    c(temperature_adaptation_terms, precipitation_adaptation_terms),
    collapse = " + "
  )
)

m_ada <- fit_climate_model(
  model_ada,
  model_data = data_adaptation
)

etable(m_ada, fitstat = ~n + r2 + wr2)

# Evaluate the fitted response at three representative lagged-income levels.
income_probabilities <- c(0.05, 0.50, 0.95)
income_percentile_names <- c("5th percentile", "Median", "95th percentile")
income_panel_rows <- est_row(data_adaptation) & valid_income_lag
income_log_levels <- unname(stats::quantile(
  data_adaptation$lag_lgrp_pc_usd_1[income_panel_rows],
  probs = income_probabilities,
  na.rm = TRUE
))

# Use the knots and boundary knots from the spline fitted in the model data.
income_spline_levels <- predict(
  income_spline_fit,
  newx = income_log_levels
)
colnames(income_spline_levels) <- spline_columns

income_scenarios <- tibble::tibble(
  income_percentile = income_percentile_names,
  log_income = income_log_levels,
  income = exp(income_log_levels),
  income_label = paste0(
    income_percentile_names,
    " ($",
    format(
      round(exp(income_log_levels)),
      big.mark = ",",
      scientific = FALSE,
      trim = TRUE
    ),
    ")"
  )
)
income_scenarios$income_label <- factor(
  income_scenarios$income_label,
  levels = income_scenarios$income_label
)

temperature_response_by_income <- adaptation_response_data(
  m_ada,
  bin_var = "TM_bin_signed",
  model_data = data_adaptation,
  spline_values = income_spline_levels,
  scenarios = income_scenarios
)
precipitation_response_by_income <- adaptation_response_data(
  m_ada,
  bin_var = "RR_bin_signed",
  model_data = data_adaptation,
  spline_values = income_spline_levels,
  scenarios = income_scenarios
)

plot_adaptation_response(
  temperature_response_by_income,
  "Temperature"
)
plot_adaptation_response(
  precipitation_response_by_income,
  "Precipitation"
)

```

## GDP growth during extreme temperature and precipitation shocks

The violins compare residual GDP-per-capita growth in years with shocks within
2.5 standard deviations of a region's moving climate baseline against the very
hot/wet and very cold/dry tails. Growth is first residualized using region and
year fixed effects plus region-specific linear year trends, matching the
`GID_1 + year + GID_1[year]` convention. Temperature and precipitation are
shown in separate panels because a region-year can belong to different tail
categories for the two climate variables.

```{r violin-gdp-growth-extreme-shocks, fig.width=10, fig.height=5.5}
shock_levels <- c(
  "Within +/-2.5 SD",
  "> 2.5 SD (very hot/wet)",
  "< -2.5 SD (very cold/dry)"
)

gdp_growth_fe_data <- data %>%
  ungroup() %>%
  filter(
    is.finite(dlgrp_pc_usd),
    !is.na(GID_1),
    !is.na(year)
  )

gdp_growth_fe_model <- fixest::feols(
  dlgrp_pc_usd ~ 1 | GID_1 + year + GID_1[year],
  data = gdp_growth_fe_data
)

gdp_growth_fe_data <- gdp_growth_fe_data %>%
  mutate(
    # Keep an NA placeholder for observations removed internally by fixest
    # (for example, fixed-effect singletons) so rows remain aligned.
    gdp_growth_residual = stats::residuals(
      gdp_growth_fe_model,
      na.rm = FALSE
    )
  )

gdp_growth_by_shock <- dplyr::bind_rows(
  gdp_growth_fe_data %>%
    transmute(
      climate_variable = "Temperature",
      shock = zTM,
      gdp_growth = gdp_growth_residual
    ),
  gdp_growth_fe_data %>%
    transmute(
      climate_variable = "Precipitation",
      shock = zRR,
      gdp_growth = gdp_growth_residual
    )
) %>%
  filter(is.finite(shock), is.finite(gdp_growth)) %>%
  mutate(
    shock_group = case_when(
      shock > 2.5 ~ "> 2.5 SD (very hot/wet)",
      shock < -2.5 ~ "< -2.5 SD (very cold/dry)",
      TRUE ~ "Within +/-2.5 SD"
    ),
    shock_group = factor(shock_group, levels = shock_levels),
    climate_variable = factor(
      climate_variable,
      levels = c("Temperature", "Precipitation")
    )
  )

shock_group_counts <- gdp_growth_by_shock %>%
  count(climate_variable, shock_group, .drop = FALSE)

shock_group_counts

ggplot(
  gdp_growth_by_shock,
  aes(x = shock_group, y = 100 * gdp_growth, fill = shock_group)
) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = "grey30",
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    fill = "white",
    alpha = 0.8,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ climate_variable) +
  scale_fill_manual(
    values = c(
      "Within +/-2.5 SD" = "grey70",
      "> 2.5 SD (very hot/wet)" = "#D55E00",
      "< -2.5 SD (very cold/dry)" = "#0072B2"
    ),
    drop = FALSE
  ) +
  labs(
    title = "Residual GDP growth by standardized climate shock",
    subtitle = paste(
      "Growth residualized by region and year fixed effects",
      "and region-specific linear trends"
    ),
    x = NULL,
    y = "Residual GDP per-capita growth (percentage points)",
    fill = "Shock category"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "none"
  )
```


```{r test-absolute}

model_demeaned <- paste(
    "(TM - mean_TM_lag_30) + (TM - mean_TM_lag_30):mean_TM_lag_30",
     "+ (RR - mean_RR_lag_30) + (RR - mean_RR_lag_30):mean_RR_lag_30",
    "+ zTM^2 + zRR^2"
)

m_dem <- fit_climate_model(
  model_demeaned
)

etable(m_dem)


model_demeaned <- paste(
    "(TM - mean_TM_all) + (TM - mean_TM_all):mean_TM_all",
     "+ (RR - mean_RR_all) + (RR - mean_RR_all):mean_RR_all",
    "+ abs_zTMp + abs_zTMm + abs_zTMp:mean_TM_all + abs_zTMm:mean_TM_all + + abs_zRRp + abs_zRRm + abs_zRRp:mean_RR_all + abs_zTMm:mean_RR_all"
)

m_dem2 <- fit_climate_model(
  model_demeaned
)

etable(m_dem,m_dem2)


```

# Test 12: Signed anomaly bins against the Hamilton trend

The signed-bin specification of Test 6, with the anomalies measured against the
Hamilton-filter trend instead of the lagged 30-year mean:
zTM = (TM - trend_TM) / sd(TM - trend_TM), where the SD is taken over each
region's entire climate series (`build_dat(baseline = "trend")`). The trend
absorbs warming, so the anomalies are centred much closer to zero. The bins keep
the same half-SD breaks and codes, so the two models are directly comparable.

```{r test-signed-bins-trend, fig.width=8, fig.height=5}
data_trend <- build_dat(
  econ_data = params$econ_data,
  config = panel_cfg,
  baseline = "trend"
)

m_bins_signed_trend <- fit_climate_model(
  model_bins_signed,
  model_data = data_trend
)

# Where the anomalies fall under each baseline.
table(lagged_30_year_mean = data$TM_bin_signed[!is.na(data$dlgrp_pc_usd)])
table(hamilton_trend = data_trend$TM_bin_signed[!is.na(data_trend$dlgrp_pc_usd)])

etable(
  m_bins_signed,
  m_bins_signed_trend,
  headers = c("Lagged 30-year mean", "Hamilton trend"),
  fitstat = ~n + r2 + wr2
)

iplot(
  list(m_bins_signed, m_bins_signed_trend),
  i.select = 1,
  main = "Signed temperature bins: lagged 30-year mean vs Hamilton trend"
)
legend("bottomleft", col = 1:2, pch = c(20, 17),
       legend = c("Lagged 30-year mean", "Hamilton trend"), bty = "n")

iplot(
  list(m_bins_signed, m_bins_signed_trend),
  i.select = 2,
  main = "Signed precipitation bins: lagged 30-year mean vs Hamilton trend"
)
legend("bottomleft", col = 1:2, pch = c(20, 17),
       legend = c("Lagged 30-year mean", "Hamilton trend"), bty = "n")

```
