---
title: "Tests of Climate Response Functions"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: show
params:
  project_setup_script: null
  climate_velocity_years: 5
  econ_data: "DOSE"
---

# Purpose

This notebook compares alternative functional forms for temperature and
precipitation effects on subnational GDP-per-capita growth. Every model uses the
same climate controls, fixed effects, panel identifiers, and clustered standard
errors so that differences can be attributed to the tested response function.

By default, the notebook loads the processed economic and climate parquet files
from `econometrics/data`. Set `project_setup_script` only when project-specific
setup is needed before those files are loaded. Set `econ_data` to `PWT`, `WDI`,
`DOSE` (DOSE v2.14); the loader selects the corresponding GADM0 or GADM1
climate files automatically.

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "arrow", "curl", "dplyr", "fixest", "ggplot2", "jsonlite", "purrr",
  "readxl", "tibble", "tidyselect", "zoo"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages)) {
  stop(
    "Install missing packages before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(dplyr)
library(fixest)
library(ggplot2)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

if (!is.null(params$project_setup_script) &&
    length(params$project_setup_script) == 1L &&
    !is.na(params$project_setup_script) &&
    nzchar(params$project_setup_script)) {
  source(params$project_setup_script, local = globalenv())
}
```

## Shared specification

The common specification contains quadratic long-run temperature and
precipitation terms. All tests add alternative weather-response terms to this
base while retaining the same fixed effects.

```{r shared-specification}
fixed_effects <- "year + GID_1[year]"
pan_id <- c("GID_1", "year")
outcome <- "dlgrp_pc_usd"

base_climate <- "TM + TM:mean_TM_all + RR + RR:mean_RR_all"
model <- paste(base_climate, "+ zRR^2 + zTM^2")

ECON_YEAR_MIN <- 1950

# Climate selection, passed straight to the Weighted Climate Dataset downloader
# (Gortan, Testa, Fagiolo and Lamperti 2024). CRU TS starts in 1901, so the
# 30-year trailing window is complete from 1930 rather than from 1970 as it is
# with ERA5's 1940 start. "concurrent population" weights each year by that
# year's population instead of by a single base year, so it carries no base
# year and `CLIMATE_WEIGHT_YEAR` is ignored.
CLIMATE_SOURCE <- "CRU TS"
CLIMATE_SOURCE_CODE <- "cru"
CLIMATE_WEIGHT <- "concurrent population"
CLIMATE_WEIGHT_YEAR <- NULL

# Fixed reference periods and Hamilton-filter settings. 
PERIOD_ALL <- c(1990L, 2019L)
PERIOD_PRE <- c(1960L, 1989L)
HAMILTON_LAGS <- 4L
HAMILTON_H_FIXED <- 2L

CLIMATE_VELOCITY_YEARS <- as.integer(params$climate_velocity_years)
CLIMATE_WINDOWS <- c(5L, 10L, 20L, 30L)

if (length(CLIMATE_VELOCITY_YEARS) != 1L ||
    is.na(CLIMATE_VELOCITY_YEARS) ||
    CLIMATE_VELOCITY_YEARS < 2L) {
  stop("params$climate_velocity_years must be an integer of at least 2.")
}

# Retain enough pre-estimation climate observations to calculate the first
# trailing climate-velocity window, as well as the six-year model lags.
CLIM_HISTORY_YEARS <- max(6L, CLIMATE_VELOCITY_YEARS - 1L)

make_formula <- function(terms) {
  as.formula(paste(outcome, "~", terms, "|", fixed_effects))
}

fit_climate_model <- function(terms, model_data = data) {
  fixest::feols(
    make_formula(terms),
    data = model_data,
    panel.id = pan_id,
    cluster = ~GID_1
  )
}

# Retained as `f` for the split-sample tests and interactive use.
f <- make_formula(model)
```

# Data preparation

The economic sample starts in `r ECON_YEAR_MIN`. Earlier climate observations
are kept as lag carriers and to calculate climate velocity at the beginning of
the estimation period.

```{r build-data}
est_row <- function(d) !is.na(d$dlgrp_pc_usd)

# Match on calendar year so that gaps produce NA rather than an incorrect lag.
lag_by_year <- function(x, year, k = 1) {
  x[match(year - k, year)]
}

# The climate series now come straight from the Weighted Climate Dataset
# through the repository's downloader, so the notebook no longer depends on a
# pre-processed climate parquet and derives the moments itself.
# `hamilton_trend()` is shared with prepare_climate_data.R so the trend operator
# is identical to the one that produced the previous inputs.
local({
  helpers <- c("download_weighted_climate_data.R", "climate_transforms.R")
  for (helper in helpers) {
    candidates <- unique(c(
      helper, file.path("..", helper), file.path("econometrics", helper)
    ))
    existing <- candidates[file.exists(candidates)]
    if (!length(existing)) {
      stop("Helper script not found: ", helper)
    }
    source(existing[[1]], local = globalenv())
  }
})

# Annual series for one geographical resolution. Temperature is the mean of the
# monthly values and precipitation their sum, which is how the dashboard
# aggregates them. Precipitation is converted from millimetres to metres so the
# level bins further down stay in metres. Downloads are cached on disk by the
# helper, so repeated calls cost nothing after the first.
load_climate_series <- function(geo_resolution) {
  is_national <- identical(geo_resolution, "gadm0")
  id_col <- if (is_national) "GID_0" else "GID_1"
  # The dashboard names gadm1 units "AFG_1_1" while GADM and the economic
  # panels use "AFG.1_1", so the first underscore becomes a dot. gadm0 ids are
  # plain ISO3 codes and need no change. Units that do not match the expected
  # shape are dropped, as in prepare_climate_data.R.
  unit_pattern <- if (is_national) "^[A-Z0-9]{3}$" else "^[A-Z0-9]{3}\\..+"

  fetch <- function(variable, value_code, value_name) {
    out <- tibble::as_tibble(wcd_get(
      variable = variable,
      source = CLIMATE_SOURCE,
      geo_resolution = geo_resolution,
      weight = CLIMATE_WEIGHT,
      weight_year = CLIMATE_WEIGHT_YEAR,
      time_frequency = "yearly",
      verbose = FALSE
    ))
    missing_cols <- setdiff(c("date", id_col, value_code), names(out))
    if (length(missing_cols)) {
      stop(
        "The downloader returned no ", paste(missing_cols, collapse = ", "),
        " column for ", variable, " at ", geo_resolution, "."
      )
    }
    out %>%
      transmute(
        year = as.integer(date),
        unit = .data[[id_col]],
        !!value_name := .data[[value_code]]
      )
  }

  inner_join(
    fetch("avg. temperature", "tmp", "TM"),
    fetch("precipitation", "pre", "RR"),
    by = c("year", "unit")
  ) %>%
    mutate(
      GID_1 = if (is_national) unit else sub("^([^_]+)_", "\\1.", unit)
    ) %>%
    filter(grepl(unit_pattern, GID_1)) %>%
    transmute(
      gadm_level = geo_resolution,
      climate_source = CLIMATE_SOURCE_CODE,
      GID_0 = substr(GID_1, 1, 3),
      GID_1,
      year,
      TM,
      RR = RR / 1000
    ) %>%
    filter(!is.na(TM) | !is.na(RR))
}

# Rolling moments are right-aligned and include the current year, exactly as in
# prepare_climate_data.R. The models lag them by one year so that a realisation
# never enters its own baseline.
add_rolling_moments <- function(unit_series) {
  for (variable in c("TM", "RR")) {
    values <- unit_series[[variable]]
    for (window in CLIMATE_WINDOWS) {
      unit_series[[paste0("mean_", variable, "_", window)]] <-
        zoo::rollmeanr(values, k = window, fill = NA_real_)
      unit_series[[paste0("sd_", variable, "_", window)]] <-
        zoo::rollapplyr(values, width = window, FUN = stats::sd, fill = NA_real_)
    }
  }
  unit_series
}

in_period <- function(year, period) {
  year >= period[[1]] & year <= period[[2]]
}

# Fixed-period moments and trends, plus the trailing rolling moments. The
# rolling operators assume a gap-free annual series within each unit, which is
# checked rather than assumed.
derive_climate_moments <- function(series) {
  gaps <- series %>%
    group_by(gadm_level, GID_1) %>%
    summarise(
      complete = n() == max(year) - min(year) + 1L,
      .groups = "drop"
    ) %>%
    filter(!complete)
  if (nrow(gaps)) {
    stop(
      "Gaps in the annual climate series for ", nrow(gaps),
      " unit(s), e.g. ", paste(utils::head(gaps$GID_1, 3), collapse = ", "),
      ". The rolling moments assume consecutive years."
    )
  }

  series %>%
    group_by(gadm_level, climate_source, GID_0, GID_1) %>%
    arrange(year, .by_group = TRUE) %>%
    group_modify(~add_rolling_moments(.x)) %>%
    mutate(
      dTM = TM - lag_by_year(TM, year),
      dRR = RR - lag_by_year(RR, year),
      mean_TM_all = mean(TM[in_period(year, PERIOD_ALL)], na.rm = TRUE),
      sd_TM_all = sd(TM[in_period(year, PERIOD_ALL)], na.rm = TRUE),
      mean_RR_all = mean(RR[in_period(year, PERIOD_ALL)], na.rm = TRUE),
      sd_RR_all = sd(RR[in_period(year, PERIOD_ALL)], na.rm = TRUE),
      mean_TM_pre = mean(TM[in_period(year, PERIOD_PRE)], na.rm = TRUE),
      sd_TM_pre = sd(TM[in_period(year, PERIOD_PRE)], na.rm = TRUE),
      mean_RR_pre = mean(RR[in_period(year, PERIOD_PRE)], na.rm = TRUE),
      sd_RR_pre = sd(RR[in_period(year, PERIOD_PRE)], na.rm = TRUE),
      trend_TM = hamilton_trend(
        TM, h = HAMILTON_H_FIXED, p = HAMILTON_LAGS
      ),
      trend_RR = hamilton_trend(
        RR, h = HAMILTON_H_FIXED, p = HAMILTON_LAGS
      )
    ) %>%
    ungroup()
}

# rmarkdown may knit from either the project root or this notebook's directory,
# so data paths are resolved against both.
data_root <- function() {
  if (dir.exists("data")) "data" else file.path("..", "data")
}

data_file <- function(name) {
  path <- file.path(data_root(), name)
  if (!file.exists(path)) {
    stop("Raw input not found: ", path)
  }
  path
}

# Derived panels are cached on disk so a session does not repeat the climate
# download, the rolling moments, or the Excel read. Each cache carries a JSON
# sidecar recording exactly the specification that produced it; when the current
# settings no longer match the sidecar the panel is rebuilt and both are
# rewritten. Deleting the cache directory is always safe.
# The sidecar records the settings that produced a cache, but settings alone do
# not pin the result: editing a builder while the constants stay put would
# otherwise leave a stale panel in place. Including the builders' own source in
# the specification makes any change to the derivation invalidate the cache.
code_fingerprint <- function(...) {
  vapply(
    list(...),
    function(f) paste(deparse(body(f)), collapse = "\n"),
    character(1)
  )
}

cached_panel <- function(tag, spec, build) {
  cache_dir <- file.path(data_root(), "cache")
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
  panel_path <- file.path(cache_dir, paste0(tag, ".parquet"))
  spec_path <- file.path(cache_dir, paste0(tag, ".json"))
  spec_json <- as.character(
    jsonlite::toJSON(spec, auto_unbox = TRUE, digits = NA, null = "null")
  )

  if (file.exists(panel_path) && file.exists(spec_path)) {
    stored <- paste(readLines(spec_path, warn = FALSE), collapse = "")
    if (identical(stored, spec_json)) {
      message("Reusing cached ", basename(panel_path))
      return(tibble::as_tibble(arrow::read_parquet(panel_path)))
    }
    message("Specification changed, rebuilding ", basename(panel_path))
  }

  panel <- build()
  arrow::write_parquet(panel, panel_path)
  writeLines(spec_json, spec_path)
  panel
}

# Economic panels are built from their raw sources here, so the notebook no
# longer needs prepare_econ_data.R to have been run. Only the two sources this
# notebook uses are covered.
ECON_RAW_FILES <- c(
  DOSE_V2_14 = "DOSE_V2.14.csv",
  PWT110 = "pwt110.xlsx"
)

read_econ_source <- function(stored_source) {
  raw_file <- data_file(ECON_RAW_FILES[[stored_source]])

  if (identical(stored_source, "DOSE_V2_14")) {
    utils::read.csv(raw_file) %>%
      select(year, GID_0, GID_1, grp_pc_usd_2015, ag_grp_pc_usd_2015, pop) %>%
      rename(grp_pc_usd = grp_pc_usd_2015) %>%
      mutate(
        share_ag_gdp = ag_grp_pc_usd_2015 / grp_pc_usd,
        econ_source = "DOSE_V2_14",
        gadm_level = "gadm1"
      ) %>%
      select(-ag_grp_pc_usd_2015) %>%
      filter(!GID_1 %in% c(" ", ""))
  } else {
    # Penn World Table 11.0. `rgdpna` and `rnna` are the constant-national-prices
    # pair, so real GDP and the capital stock are deflated the same way and their
    # ratio is meaningful over time; the chained-PPP series are the ones for
    # cross-country LEVEL comparisons, which is not what this panel is used for.
    # Both are millions of 2021 US$ against millions of people, so the ratios are
    # 2021 US$ per person -- a different base year from DOSE's 2015 US$, which is
    # fine because the sources are only ever compared through growth rates.
    readxl::read_excel(raw_file, sheet = "Data") %>%
      mutate(across(c(rgdpna, rnna, pop, emp), as.numeric)) %>%
      transmute(
        year = as.integer(year),
        GID_0 = as.character(countrycode),
        GID_1 = as.character(countrycode),
        grp_pc_usd = rgdpna / pop,      # Real GDP per capita, 2021 US$.
        k_pc_usd = rnna / pop,          # Capital stock per capita, 2021 US$.
        share_emp_pop = emp / pop,      # Persons engaged per head.
        share_ag_gdp = NA_real_,
        econ_source = "PWT110",
        gadm_level = "gadm0"
      )
  }
}

# Growth rates are matched on calendar year, so a gap in a unit's series yields
# NA rather than a growth rate silently spanning the gap. prepare_econ_data.R
# used a positional lag here, which is the one substantive difference between
# this builder and that script.
build_econ_panel <- function(stored_source) {
  read_econ_source(stored_source) %>%
    filter(!is.na(grp_pc_usd), grp_pc_usd > 0) %>%
    group_by(GID_1, gadm_level, econ_source) %>%
    arrange(year, .by_group = TRUE) %>%
    mutate(
      lgrp_pc_usd = log(grp_pc_usd),
      lag_lgrp_pc_usd = lag_by_year(lgrp_pc_usd, year),
      dlgrp_pc_usd = lgrp_pc_usd - lag_lgrp_pc_usd,
      dg = grp_pc_usd / lag_by_year(grp_pc_usd, year) - 1,
      lag_share_ag_gdp = lag_by_year(share_ag_gdp, year)
    ) %>%
    mutate(lag_dlgrp_pc_usd = lag_by_year(dlgrp_pc_usd, year)) %>%
    ungroup() %>%
    mutate(GID_0 = dplyr::coalesce(GID_0, substr(GID_1, 1, 3)))
}

load_econ_panel <- function(stored_source) {
  cached_panel(
    paste0("econ_", tolower(stored_source)),
    list(
      econ_source = stored_source,
      raw_file = unname(ECON_RAW_FILES[[stored_source]]),
      code = code_fingerprint(read_econ_source, build_econ_panel)
    ),
    function() build_econ_panel(stored_source)
  )
}

load_climate_panel <- function(geo_resolution) {
  cached_panel(
    paste0(
      "climate_", geo_resolution, "_", CLIMATE_SOURCE_CODE, "_",
      gsub("[^a-z0-9]+", "_", tolower(CLIMATE_WEIGHT))
    ),
    list(
      geo_resolution = geo_resolution,
      climate_source = CLIMATE_SOURCE,
      weight = CLIMATE_WEIGHT,
      weight_year = if (is.null(CLIMATE_WEIGHT_YEAR)) "none" else CLIMATE_WEIGHT_YEAR,
      windows = CLIMATE_WINDOWS,
      period_all = PERIOD_ALL,
      period_pre = PERIOD_PRE,
      hamilton_h = HAMILTON_H_FIXED,
      hamilton_p = HAMILTON_LAGS,
      code = code_fingerprint(
        load_climate_series, add_rolling_moments, derive_climate_moments
      )
    ),
    function() derive_climate_moments(load_climate_series(geo_resolution))
  )
}

# Both halves of the panel are built from raw sources and cached, so a fresh
# checkout needs no preprocessing script: the economic series come from the DOSE
# csv or the PWT workbook, the climate series from the Weighted Climate Dataset.
build_dat <- function(
    econ_data = c("DOSE", "DOSE_V2_14", "PWT", "PWT110"),
    climate_series = NULL
) {
  econ_data <- match.arg(econ_data)

  # "DOSE" and "PWT" are version-agnostic aliases, so a version bump is a change
  # to ECON_RAW_FILES and to the alias target rather than to every caller.
  econ_source_aliases <- c(
    PWT = "PWT110",
    PWT110 = "PWT110",
    DOSE = "DOSE_V2_14",
    DOSE_V2_14 = "DOSE_V2_14"
  )
  gadm_level_by_source <- c(
    PWT = "gadm0",
    PWT110 = "gadm0",
    DOSE = "gadm1",
    DOSE_V2_14 = "gadm1"
  )

  stored_econ_source <- unname(econ_source_aliases[[econ_data]])
  target_gadm_level <- unname(gadm_level_by_source[[econ_data]])

  econ <- load_econ_panel(stored_econ_source) %>%
    select(
      year, GID_0, GID_1, grp_pc_usd, lgrp_pc_usd, dlgrp_pc_usd,
      econ_source, gadm_level
    ) %>%
    distinct() %>%
    filter(
      gadm_level == target_gadm_level,
      year >= ECON_YEAR_MIN,
      !is.na(dlgrp_pc_usd)
    )

  if (!nrow(econ)) {
    stop(
      "No ", target_gadm_level, " economic observations found for econ_data = '",
      econ_data, "' (stored source '", stored_econ_source, "')."
    )
  }

  # `climate_series` lets a caller pass an already derived panel, which bypasses
  # the cache entirely when experimenting with alternative moments.
  climate <- climate_series
  if (is.null(climate)) {
    climate <- load_climate_panel(target_gadm_level)
  }
  climate <- climate %>% filter(gadm_level == target_gadm_level)

  if (!nrow(climate)) {
    stop(
      "No ", target_gadm_level, " climate observations for econ_data = '",
      econ_data, "'. Check CLIMATE_SOURCE and CLIMATE_WEIGHT."
    )
  }

  # The moments are already computed over the full series, so trimming the
  # history here only drops rows that carry no lag any model needs.
  climate %>%
    filter(year >= ECON_YEAR_MIN - CLIM_HISTORY_YEARS) %>%
    left_join(
      econ,
      by = c("gadm_level", "GID_0", "GID_1", "year")
    ) %>%
    group_by(gadm_level, GID_1) %>%
    filter(any(!is.na(dlgrp_pc_usd))) %>%
    filter(year <= max(year[!is.na(dlgrp_pc_usd)])) %>%
    mutate(
      log_gdp_av = log(mean(grp_pc_usd, na.rm = TRUE)),
      mean_TM_sample = mean(TM[!is.na(dlgrp_pc_usd)], na.rm = TRUE),
      mean_RR_sample = mean(RR[!is.na(dlgrp_pc_usd)], na.rm = TRUE)
    ) %>%
    arrange(year, .by_group = TRUE) %>%
    mutate(
      mean_TM_lag_30 = lag_by_year(mean_TM_30, year),
      mean_RR_lag_30 = lag_by_year(mean_RR_30, year),
      mean_TM_lag_20 = lag_by_year(mean_TM_20, year),
      mean_RR_lag_20 = lag_by_year(mean_RR_20, year),
      mean_TM_lag_10 = lag_by_year(mean_TM_10, year),
      mean_RR_lag_10 = lag_by_year(mean_RR_10, year),
      mean_TM_lag_5 = lag_by_year(mean_TM_5, year),
      mean_RR_lag_5 = lag_by_year(mean_RR_5, year),
      sd_TM_lag_30 = lag_by_year(sd_TM_30, year),
      sd_RR_lag_30 = lag_by_year(sd_RR_30, year),
      sd_TM_lag_20 = lag_by_year(sd_TM_20, year),
      sd_RR_lag_20 = lag_by_year(sd_RR_20, year),
      sd_TM_lag_10 = lag_by_year(sd_TM_10, year),
      sd_RR_lag_10 = lag_by_year(sd_RR_10, year),
      sd_TM_lag_5 = lag_by_year(sd_TM_5, year),
      sd_RR_lag_5 = lag_by_year(sd_RR_5, year),
    ) %>%
    ungroup() %>%
  mutate(
    # Center the cross-sectional adaptation covariates so that the main bin
    # coefficients describe a region at the estimation-sample means.
    log_gdp_av_c = log_gdp_av - mean(
      log_gdp_av[!is.na(dlgrp_pc_usd)], na.rm = TRUE
    ),
    mean_TM_sample_c = mean_TM_sample - mean(
      mean_TM_sample[!is.na(dlgrp_pc_usd)], na.rm = TRUE
    ),
    mean_RR_sample_c = mean_RR_sample - mean(
      mean_RR_sample[!is.na(dlgrp_pc_usd)], na.rm = TRUE
    ),
    mean_TM = mean_TM_lag_30,
    mean_RR = mean_RR_lag_30,
    zTM = (TM - mean_TM_lag_30) / sd_TM_lag_30,
    zRR = (RR - mean_RR_lag_30) / sd_RR_lag_30,
    abs_zTM = abs(zTM),
    abs_zRR = abs(zRR),
    # Magnitudes of positive (+) and negative (-) standardized deviations.
    abs_zTMp = pmax(zTM, 0),
    abs_zTMm = pmax(-zTM, 0),
    abs_zRRp = pmax(zRR, 0),
    abs_zRRm = pmax(-zRR, 0),

    TM_0_05 = pmin(abs_zTM, 0.5),
    TM_05_15 = pmin(pmax(abs_zTM - 0.5, 0), 1.5),
    TM_15_25 = pmin(pmax(abs_zTM - 1.5, 0), 2.5),
    TM_25p = pmax(abs_zTM - 2.5, 0),

    RR_0_05 = pmin(abs_zRR, 0.5),
    RR_05_15 = pmin(pmax(abs_zRR - 0.5, 0), 1.5),
    RR_15_25 = pmin(pmax(abs_zRR - 0.5, 0), 2.5),
    RR_25p = pmax(abs_zRR - 2.5, 0),

    TM_bin = cut(
      abs_zTM,
      breaks = c(0, 1, 2, Inf),
      labels = c("0-1", "1-2", ">2"),
      include.lowest = TRUE,
      right = FALSE
    ),
    RR_bin = cut(
      abs_zRR,
      breaks = c(0, 1, 2, Inf),
      labels = c("0-1", "1-2", ">2"),
      include.lowest = TRUE,
      right = FALSE
    ),
    TM_bin_fine = cut(
      abs_zTM,
      breaks = c(0, 0.5, 1, 1.5, 2, 2.5, Inf),
      labels = c(
        "0-.5", ".5-1", "1-1.5", "1.5-2",
        "2-2.5", ">2.5"
      ),
      include.lowest = TRUE,
      right = FALSE
    ),
    RR_bin_fine = cut(
      abs_zRR,
      breaks = c(0, 0.5, 1, 1.5, 2, 2.5, Inf),
      labels = c(
        "0-.5", ".5-1", "1-1.5", "1.5-2",
        "2-2.5", ">2.5"
      ),
      include.lowest = TRUE,
      right = FALSE
    ),
    TM_bin_level = cut(
      TM,
      breaks = c(-Inf, 0, 5, 10, 15, 20, 25, Inf),
      labels = c(
        "<0", "0-5", "5-10", "10-15", "15-20", "20-25", ">=25"
      ),
      right = FALSE
    ),
    RR_bin_level = cut(
      RR,
      breaks = c(-Inf, 0.5, 1, 1.5, 2, 2.5, Inf),
      labels = c(
        "<0.5", "0.5-1", "1-1.5", "1.5-2", "2-2.5", ">=2.5"
      ),
      right = FALSE
    )
  ) %>%
  mutate(
    TM_bin_signed = case_when(
      zTM < -2.5 ~ -2.75,
      zTM < -2.0 ~ -2.25,
      zTM < -1.5 ~ -1.75,
      zTM < -1.0 ~ -1.25,
      zTM < -0.5 ~ -0.75,
      zTM <  0.5 ~  0.00,
      zTM <  1.0 ~  0.75,
      zTM <  1.5 ~  1.25,
      zTM <  2.0 ~  1.75,
      zTM <  2.5 ~  2.25,
      zTM >= 2.5 ~  2.75
    ),
    RR_bin_signed = case_when(
      zRR < -2.5 ~ -2.75,
      zRR < -2.0 ~ -2.25,
      zRR < -1.5 ~ -1.75,
      zRR < -1.0 ~ -1.25,
      zRR < -0.5 ~ -0.75,
      zRR <  0.5 ~  0.00,
      zRR <  1.0 ~  0.75,
      zRR <  1.5 ~  1.25,
      zRR <  2.0 ~  1.75,
      zRR <  2.5 ~  2.25,
      zRR >= 2.5 ~  2.75
    )
  )
}


data <- build_dat(econ_data = params$econ_data)

data_iso3 <- build_dat(econ_data = "PWT")

```

## Weather-response variables

Weather deviations are standardized by each region's long-run variability.
The derived variables support the piecewise-linear and discrete-bin tests below.

```{r derive-weather-variables}
table(data$TM_bin_signed, useNA = "ifany")
table(data$RR_bin_signed, useNA = "ifany")
table(data$TM_bin_level, useNA = "ifany")
table(data$RR_bin_level, useNA = "ifany")
table(data_iso3$TM_bin_signed, useNA = "ifany")
table(data_iso3$RR_bin_signed, useNA = "ifany")
```

# Test: Correlation between fine temperature and precipitation bins

The Spearman test treats the fine bins as ordered categories. The bin-pair
matrix reports correlations between the indicator for each temperature bin and
the indicator for each precipitation bin.

```{r test-fine-bin-correlation}
fine_bin_data <- data %>%
  filter(
    !is.na(dlgrp_pc_usd),
    !is.na(TM_bin_fine),
    !is.na(RR_bin_fine)
  ) %>%
  transmute(
    TM_bin_fine,
    RR_bin_fine,
    TM_bin_rank = as.integer(TM_bin_fine),
    RR_bin_rank = as.integer(RR_bin_fine)
  )

fine_bin_correlation_test <- cor.test(
  fine_bin_data$TM_bin_rank,
  fine_bin_data$RR_bin_rank,
  method = "spearman",
  exact = FALSE
)

fine_bin_correlation_test

TM_bin_indicators <- model.matrix(
  ~ TM_bin_fine - 1,
  data = fine_bin_data
)
RR_bin_indicators <- model.matrix(
  ~ RR_bin_fine - 1,
  data = fine_bin_data
)

colnames(TM_bin_indicators) <- paste0(
  "TM_",
  levels(fine_bin_data$TM_bin_fine)
)
colnames(RR_bin_indicators) <- paste0(
  "RR_",
  levels(fine_bin_data$RR_bin_fine)
)

fine_bin_pair_correlations <- cor(
  TM_bin_indicators,
  RR_bin_indicators
)

round(fine_bin_pair_correlations, 3)
```


# Test 1: Quadratic weather deviations

This benchmark asks whether squared standardized temperature and precipitation
deviations add information beyond the main BHM climate specification. The two
models use the same estimation sample. A clustered joint Wald test evaluates
the null that both added deviation coefficients are zero.

IDEAS:
- transform the dev variable to z-zmean/(2*zsd) to weight less lower deviation as they show no statistical effects in bins
- main specification should be this + distributed lag models of past SDs?
- check last chatGPT chat for suggestions on spells

```{r test-absolute}

model_bhm <- "TM + TM:mean_TM_lag_30 + RR + RR:mean_RR_lag_30"

f_bhm <- make_formula(model_bhm)
m_bhm <- fit_climate_model(
  model_bhm
)

model_base <- paste(
  "TM + TM:mean_TM_lag_30 + RR + RR:mean_RR_lag_30",
  "+ abs_zTMp + abs_zTMm + abs_zRRp + abs_zRRm + mean_TM_lag_30 + mean_RR_lag_30"
)

f_base <- make_formula(model_base)
m_base <- fit_climate_model(
  model_base
)

m_base_iso3 <- fit_climate_model(
  model_base,
  model_data = data_iso3
)

deviation_information_test <- fixest::wald(
  m_base,
  keep = "^z(TM|RR^2)$",
  print = FALSE
)

etable(
  m_base_iso3,
  m_bhm,
  m_base,
  headers = c("Iso3", "Naive", "Full"),
  fitstat = ~n + r2 + wr2 + aic + bic
)

iplot(m_base)
```



```{r test-quadratic}
model_abs <- paste(
  base_climate,
  "+ abs(zTM) + abs(zRR)"
)
m_abs <- fit_climate_model(model_abs)

etable(m_base,m_abs, fitstat = ~n + r2 + wr2)

```


# Test 1: lags

How much memory is in the system?

```{r test lags}

data <- data |> group_by(GID_1) |> 
  mutate(dTM=TM-lag_by_year(TM,year),dRR=RR-lag_by_year(RR,year))

model_lags <- paste(
#  "l(TM,0:5) + l(TM,0:5):mean_TM_all",
#  "+ l(RR,0:5) + l(RR,0:5):mean_RR_all",
  "+ l(zRR^2,0:10) + l(zTM^2,0:10)"
)

f_lags <- make_formula(model_lags)
m_lags <- fit_climate_model(model_lags)

etable(m_lags, fitstat = ~n + r2 + wr2)
  

```

```{r plot-distributed-lags, fig.width=10, fig.height=6}
# Plot both each lag coefficient and the cumulative response through that lag.
# The cumulative confidence intervals retain the cross-lag covariances from the
# GID_1-clustered variance-covariance matrix used to estimate m_lags. Terms and
# lag ranges are discovered from the fitted model, so this chunk does not need
# to be updated when model_lags changes.
lag_confidence_level <- 0.95
lag_z_critical <- qnorm(1 - (1 - lag_confidence_level) / 2)

lag_coefficients <- coef(m_lags)
lag_vcov <- vcov(m_lags)
lag_coefficient_names <- names(lag_coefficients)
lag_coefficient_names_normalized <- gsub(
  "[[:space:]]",
  "",
  lag_coefficient_names
)

lag_name_pattern <- "^l\\((.*),([0-9]+)\\)$"
lagged_positions <- grep(
  lag_name_pattern,
  lag_coefficient_names_normalized
)

if (!length(lagged_positions)) {
  stop("m_lags contains no coefficients named l(term, lag).")
}

lag_term_index <- tibble::tibble(
  expression = sub(
    lag_name_pattern,
    "\\1",
    lag_coefficient_names_normalized[lagged_positions]
  ),
  lag = as.integer(sub(
    lag_name_pattern,
    "\\2",
    lag_coefficient_names_normalized[lagged_positions]
  )),
  term = lag_coefficient_names[lagged_positions]
)

distributed_lag_data <- purrr::map_dfr(
  unique(lag_term_index$expression),
  function(lag_expression) {
    lag_terms <- lag_term_index %>%
      filter(expression == lag_expression) %>%
      arrange(lag)

    # Also support specifications written as term + l(term, 1:k), where fixest
    # prints the contemporaneous coefficient without the l(., 0) wrapper.
    if (!0L %in% lag_terms$lag) {
      lag_zero_position <- match(
        lag_expression,
        lag_coefficient_names_normalized
      )
      if (!is.na(lag_zero_position)) {
        lag_terms <- bind_rows(
          tibble::tibble(
            expression = lag_expression,
            lag = 0L,
            term = lag_coefficient_names[lag_zero_position]
          ),
          lag_terms
        ) %>%
          arrange(lag)
      }
    }

    if (anyDuplicated(lag_terms$lag)) {
      stop("Duplicate coefficients found for ", lag_expression, ".")
    }

    purrr::map_dfr(lag_terms$lag, function(current_lag) {
      current_term <- lag_terms$term[lag_terms$lag == current_lag]
      cumulative_terms <- lag_terms$term[lag_terms$lag <= current_lag]

      current_effect <- unname(lag_coefficients[current_term])
      current_se <- sqrt(max(
        as.numeric(lag_vcov[current_term, current_term]),
        0
      ))
      cumulative_effect <- sum(lag_coefficients[cumulative_terms])
      cumulative_se <- sqrt(max(
        sum(lag_vcov[cumulative_terms, cumulative_terms, drop = FALSE]),
        0
      ))

      tibble::tibble(
        expression = lag_expression,
        lag = current_lag,
        horizon = c("Effect at lag", "Cumulative through lag"),
        effect = c(current_effect, cumulative_effect),
        standard_error = c(current_se, cumulative_se)
      )
    })
  }
) %>%
  mutate(
    expression = factor(
      expression,
      levels = unique(lag_term_index$expression)
    ),
    horizon = factor(
      horizon,
      levels = c("Effect at lag", "Cumulative through lag")
    ),
    lower = effect - lag_z_critical * standard_error,
    upper = effect + lag_z_critical * standard_error
  )

ggplot(distributed_lag_data, aes(x = lag, y = effect)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_grid(expression ~ horizon, scales = "free_y") +
  scale_x_continuous(breaks = sort(unique(distributed_lag_data$lag))) +
  labs(
    x = "Lag (years)",
    y = "Effect on log GDP per-capita growth",
    title = "Distributed-lag coefficient paths",
    caption = paste0(
      "Ribbons = ", round(100 * lag_confidence_level),
      "% clustered CI; cumulative intervals include cross-lag covariance"
    )
  ) +
  theme_classic()
```

```{r test windows}

model_windows <- paste(
  base_climate,
  " + abs((TM - mean_TM_lag_30) / sd_TM_lag_30)   + abs((TM - mean_TM_lag_5) / sd_TM_lag_5)",
  " + abs((RR - mean_RR_lag_30) / sd_RR_lag_30)   + abs((RR - mean_RR_lag_5) / sd_RR_lag_5)"
)

f_windows <- make_formula(model_windows)
m_windows <- fit_climate_model(model_windows)

etable(m_windows, fitstat = ~n + r2 + wr2)
  

```


# Test 2: Piecewise-linear absolute deviations

This test allows the marginal effect of an absolute standardized anomaly to
change as it crosses increasingly extreme ranges.

```{r test-piecewise}
model_piece <- paste(
  base_climate,
  "+ TM_0_05 + TM_05_15 + TM_15_25 + TM_25p",
  "+ RR_0_05 + RR_05_15 + RR_15_25 + RR_25p"
)
f_piece <- make_formula(model_piece)
m_piece <- fit_climate_model(model_piece)

etable(m_piece, fitstat = ~n + r2 + wr2)
```

# Test 3: Coarse unsigned anomaly bins

This nonparametric specification groups absolute anomalies into three broad
ranges, allowing moderate and extreme weather to have distinct effects.

```{r test-coarse-bins}
model_bin <- paste(base_climate, "+ TM_bin + RR_bin")
f_bin <- make_formula(model_bin)
m_bin <- fit_climate_model(model_bin)

etable(m_bin, fitstat = ~n + r2 + wr2)
```

# Test 4: Fine unsigned anomaly bins

This test increases resolution to half-standard-deviation intervals and uses
the 0--0.5 standard-deviation category as the reference for both variables.

```{r test-fine-bins}
model_bins <- paste(
  base_climate,
  "+ i(TM_bin_fine, ref = \"0-.5\")",
  "+ i(RR_bin_fine, ref = \"0-.5\")"
)
f_bins <- make_formula(model_bins)
m_bins <- fit_climate_model(model_bins)

etable(m_bins, fitstat = ~n + r2 + wr2)
```


```{r moving averages interaction}

model_moving <- paste(
  "TM + TM:mean_TM_lag_30 + RR + RR:mean_RR_lag_30",
  "+ zRR^2 + zTM^2"
)

m_move <- fit_climate_model(model_moving)

etable(m_bhm,m_base,m_move)


```



## Functional-form comparison

The table places the principal alternatives side by side on their common
fixed-effect specification.

```{r compare-functional-forms}
etable(
  m_base,
  m_piece,
  m_bin,
  m_bins,
  headers = c(
    "Quadratic",
    "Piecewise linear",
    "Coarse bins",
    "Fine bins"
  ),
  fitstat = ~n + r2 + wr2
)
```

# Test 5: Variable-specific fine bins

Estimating the fine-bin response for one weather variable at a time makes its
shape easier to inspect without simultaneously estimating the other bin set.

```{r test-variable-specific-fine-bins, fig.width=8, fig.height=5}
model_bins_fine_TM <- paste(
  base_climate,
  "+ i(TM_bin_fine, ref = \"0-.5\")"
)
f_bins_fine_TM <- make_formula(model_bins_fine_TM)
m_bins_fine_TM <- fit_climate_model(model_bins_fine_TM)

model_bins_fine_RR <- paste(
  base_climate,
  "+ i(RR_bin_fine, ref = \"0-.5\")"
)
f_bins_fine_RR <- make_formula(model_bins_fine_RR)
m_bins_fine_RR <- fit_climate_model(model_bins_fine_RR)

iplot(m_bins_fine_TM, main = "Temperature anomaly bins")
iplot(m_bins_fine_RR, main = "Precipitation anomaly bins")
```

# Test 6: Signed anomaly bins

Unsigned bins impose the same response on positive and negative deviations.
This test instead distinguishes unusually cold from hot years and unusually dry
from wet years, with deviations between -0.5 and 0.5 standard deviations as the
reference category.

```{r test-signed-bins, fig.width=8, fig.height=5}
model_bins_TM_signed <- paste(
  base_climate,
  "+ i(TM_bin_signed, ref = 0)"
)
f_bins_TM_signed <- make_formula(model_bins_TM_signed)
m_bins_TM_signed <- fit_climate_model(model_bins_TM_signed)

model_bins_RR_signed <- paste(
  base_climate,
  "+ i(RR_bin_signed, ref = 0)"
)
f_bins_RR_signed <- make_formula(model_bins_RR_signed)
m_bins_RR_signed <- fit_climate_model(model_bins_RR_signed)

model_bins_signed <- paste(
    base_climate,
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
)
f_bins_signed <- make_formula(model_bins_signed)
m_bins_signed <- fit_climate_model(model_bins_signed)

iplot(m_bins_TM_signed, main = "Signed temperature anomaly bins")
iplot(m_bins_RR_signed, main = "Signed precipitation anomaly bins")

iplot(m_bins_signed, main = "Signed temperature anomaly bins")
iplot(m_bins_signed, i.select=2,main = "Signed precipitation anomaly bins")

etable(
  m_bins_TM_signed,
  m_bins_RR_signed,
  m_bins_signed,
  headers = c("Temperature", "Precipitation", "Joint"),
  fitstat = ~n + r2 + wr2
)
```

## Temperature and precipitation level bins with signed deviations

This specification adds nonparametric bins for annual temperature and
precipitation levels to the joint signed-deviation model. Temperature bins use
5-degree Celsius intervals and precipitation bins use 0.5-metre intervals. The
most populated central intervals (10--15 degrees Celsius and 0.5--1 metre) are
the level-bin references; the signed-deviation reference remains -0.5 to 0.5
within-region standard deviations. Consequently, every plotted coefficient is
conditional on the common BHM climate controls and on the other three bin sets.

```{r test-level-and-signed-bins, fig.width=8, fig.height=5, fig.show='hold', out.width='50%'}
model_bins_level_signed <- paste(
  "+ i(TM_bin_level, ref = \"10-15\")",
  "+ i(RR_bin_level, ref = \"0.5-1\")",
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
)
f_bins_level_signed <- make_formula(model_bins_level_signed)
m_bins_level_signed <- fit_climate_model(model_bins_level_signed)

iplot(
  m_bins_level_signed,
  i.select = 1,
  main = "Annual temperature bins",
  xlab = "Annual mean temperature (degrees Celsius)"
)
iplot(
  m_bins_level_signed,
  i.select = 2,
  main = "Annual precipitation bins",
  xlab = "Annual precipitation (metres)"
)
iplot(
  m_bins_level_signed,
  i.select = 3,
  main = "Signed temperature-deviation bins",
  xlab = "Temperature deviation (standard deviations)"
)
iplot(
  m_bins_level_signed,
  i.select = 4,
  main = "Signed precipitation-deviation bins",
  xlab = "Precipitation deviation (standard deviations)"
)

etable(
  m_bins_signed,
  m_bins_level_signed,
  headers = c("Signed deviations", "Level and signed bins"),
  fitstat = ~n + r2 + wr2
)
```

### Carleton-style heterogeneity by income and long-run climate

Following the heterogeneous-response construction used by Carleton and
coauthors, each annual temperature-bin indicator is interacted linearly with
the region's sample-average log income and sample-average temperature. Each
precipitation-bin indicator is analogously interacted with sample-average log
income and sample-average precipitation. The modifiers are centered over the
estimation sample, so the uninteracted bin coefficients apply at the sample
means. Region fixed effects absorb the modifiers' main effects.

The figures divide regions into terciles of income and long-run climate, then
evaluate the estimated response at the mean income and climate within each of
the nine cells. All interactions are estimated jointly; the signed anomaly bins
remain in the model as additive deviation controls.

```{r test-carleton-bin-heterogeneity, fig.width=12, fig.height=8}
model_bins_level_signed_heterogeneous <- paste(
  model_bins_level_signed,
  "+ i(TM_bin_level, log_gdp_av_c, ref = \"10-15\")",
  "+ i(TM_bin_level, mean_TM_sample_c, ref = \"10-15\")",
  "+ i(RR_bin_level, log_gdp_av_c, ref = \"0.5-1\")",
  "+ i(RR_bin_level, mean_RR_sample_c, ref = \"0.5-1\")"
)
f_bins_level_signed_heterogeneous <- make_formula(
  model_bins_level_signed_heterogeneous
)
m_bins_level_signed_heterogeneous <- fit_climate_model(
  model_bins_level_signed_heterogeneous
)

make_carleton_scenarios <- function(
    model_data,
    climate_var,
    climate_centered_var,
    climate_group_labels
) {
  income_group_labels <- c("Low income", "Middle income", "High income")

  model_data %>%
    ungroup() %>%
    filter(!is.na(dlgrp_pc_usd)) %>%
    distinct(
      GID_1,
      log_gdp_av,
      log_gdp_av_c,
      .data[[climate_var]],
      .data[[climate_centered_var]]
    ) %>%
    mutate(
      income_group = factor(
        income_group_labels[dplyr::ntile(log_gdp_av, 3L)],
        levels = income_group_labels
      ),
      climate_group = factor(
        climate_group_labels[dplyr::ntile(.data[[climate_var]], 3L)],
        levels = climate_group_labels
      )
    ) %>%
    group_by(income_group, climate_group) %>%
    summarise(
      income_value = mean(log_gdp_av_c),
      climate_value = mean(.data[[climate_centered_var]]),
      regions = n(),
      .groups = "drop"
    )
}

carleton_bin_response <- function(
    fitted_model,
    model_data,
    bin_var,
    reference_bin,
    income_var,
    climate_var,
    scenarios
) {
  model_coef <- stats::coef(fitted_model)
  model_vcov <- stats::vcov(fitted_model)
  coefficient_names <- names(model_coef)
  bin_levels <- levels(droplevels(model_data[[bin_var]]))

  purrr::map_dfr(seq_len(nrow(scenarios)), function(scenario_id) {
    scenario <- scenarios[scenario_id, ]

    purrr::map_dfr(bin_levels, function(bin_level) {
      if (identical(bin_level, reference_bin)) {
        return(tibble::tibble(
          bin = bin_level,
          estimate = 0,
          std_error = 0
        ))
      }

      main_coefficient <- paste0(bin_var, "::", bin_level)
      income_coefficient <- paste0(
        main_coefficient, ":", income_var
      )
      climate_coefficient <- paste0(
        main_coefficient, ":", climate_var
      )
      required_coefficients <- c(
        main_coefficient,
        income_coefficient,
        climate_coefficient
      )
      missing_coefficients <- setdiff(
        required_coefficients,
        coefficient_names
      )

      if (length(missing_coefficients)) {
        stop(
          "Heterogeneous bin coefficients missing from the model: ",
          paste(missing_coefficients, collapse = ", ")
        )
      }

      contrast <- setNames(numeric(length(model_coef)), coefficient_names)
      contrast[[main_coefficient]] <- 1
      contrast[[income_coefficient]] <- scenario$income_value[[1]]
      contrast[[climate_coefficient]] <- scenario$climate_value[[1]]

      estimate <- sum(contrast * model_coef)
      variance <- drop(t(contrast) %*% model_vcov %*% contrast)

      tibble::tibble(
        bin = bin_level,
        estimate = estimate,
        std_error = sqrt(max(variance, 0))
      )
    }) %>%
      mutate(
        income_group = scenario$income_group[[1]],
        climate_group = scenario$climate_group[[1]],
        regions = scenario$regions[[1]]
      )
  }) %>%
    mutate(
      bin = factor(bin, levels = bin_levels),
      conf_low = estimate - stats::qnorm(0.975) * std_error,
      conf_high = estimate + stats::qnorm(0.975) * std_error
    )
}

plot_carleton_response <- function(plot_data, title, x_label) {
  ggplot(plot_data, aes(x = bin, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, color = "grey55", linewidth = 0.4) +
    geom_line(linewidth = 0.55, color = "#0072B2") +
    geom_errorbar(
      aes(ymin = conf_low, ymax = conf_high),
      width = 0.18,
      linewidth = 0.45,
      color = "#0072B2"
    ) +
    geom_point(size = 1.8, color = "#0072B2") +
    facet_grid(income_group ~ climate_group) +
    labs(
      title = title,
      subtitle = paste(
        "Columns: long-run climate tercile; rows: sample-average income tercile"
      ),
      x = x_label,
      y = "Estimated effect relative to reference bin"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "grey92", color = "grey70")
    )
}

temperature_scenarios <- make_carleton_scenarios(
  data,
  climate_var = "mean_TM_sample",
  climate_centered_var = "mean_TM_sample_c",
  climate_group_labels = c("Cold", "Temperate", "Hot")
)
precipitation_scenarios <- make_carleton_scenarios(
  data,
  climate_var = "mean_RR_sample",
  climate_centered_var = "mean_RR_sample_c",
  climate_group_labels = c("Dry", "Moderate", "Wet")
)

temperature_carleton_response <- carleton_bin_response(
  m_bins_level_signed_heterogeneous,
  model_data = data,
  bin_var = "TM_bin_level",
  reference_bin = "10-15",
  income_var = "log_gdp_av_c",
  climate_var = "mean_TM_sample_c",
  scenarios = temperature_scenarios
)
precipitation_carleton_response <- carleton_bin_response(
  m_bins_level_signed_heterogeneous,
  model_data = data,
  bin_var = "RR_bin_level",
  reference_bin = "0.5-1",
  income_var = "log_gdp_av_c",
  climate_var = "mean_RR_sample_c",
  scenarios = precipitation_scenarios
)

plot_carleton_response(
  temperature_carleton_response,
  "Temperature-bin response by average income and temperature",
  "Annual mean temperature (degrees Celsius)"
)
plot_carleton_response(
  precipitation_carleton_response,
  "Precipitation-bin response by average income and precipitation",
  "Annual precipitation (metres)"
)

etable(
  m_bins_level_signed,
  m_bins_level_signed_heterogeneous,
  headers = c("Additive bins", "Income-climate interactions"),
  fitstat = ~n + r2 + wr2
)
```

# PREFERRED SPECIFICATION:

```{r test-double deviation}

data <- data |> 
    mutate(zzTM = pmax(1.5,abs((TM - mean_TM_lag_30) / (sd_TM_lag_30)))-1.5 ) 

m_pref <- fit_climate_model(paste(base_climate,"+ zzTM^2 + abs_zRRp"))

etable(m_base, m_pref, fitstat = ~n + r2 + wr2)

```


```{r signed, national}

m_bins_signed_iso3 <- fit_climate_model(model_bins_signed,model_data=data_iso3)
m_quad_iso3 <- fit_climate_model(model_base,model_data=data_iso3)


iplot(m_bins_signed, m_bins_signed_iso3, main = "Signed temperature anomaly bins")

```


# signed, divided by mean temperature and precip
```{r temperature and precip robust}
data_income <- data %>%
  group_by(GID_0) %>%
  mutate(av_y = mean(grp_pc_usd, na.rm = TRUE)) %>%
  ungroup()

income_cutoffs <- quantile(
  data_income$av_y,
  c(0.25, 0.5, 0.75),
  na.rm = TRUE
)
mean_TM_cutoff <- median(data_income$mean_TM_all, na.rm = TRUE)
mean_RR_cutoff <- median(data_income$mean_RR_all, na.rm = TRUE)

data_signed_groups <- data_income %>%
  mutate(
    mean_temperature_group = factor(
      if_else(mean_TM_all < mean_TM_cutoff, "Cold", "Hot"),
      levels = c("Cold", "Hot")
    ),
    mean_precipitation_group = factor(
      if_else(mean_RR_all < mean_RR_cutoff, "Dry", "Wet"),
      levels = c("Dry", "Wet")
    ),
    income_group = factor(
      if_else(av_y <= income_cutoffs[[1]], "Poor", "Rich"),
      levels = c("Poor", "Rich")
    )
  )

m_signed_tlo <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_temperature_group == "Cold"),
  panel.id = pan_id,
  vcov = ~GID_1
)


f_bins_signed_int <- make_formula(paste(
    base_climate,
  "+ i(TM_bin_signed, mean_temperature_group, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
))


m_signed_plo <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_precipitation_group == "Dry"),
  panel.id = pan_id,
  vcov = ~GID_1
)


m_signed_phi <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(mean_precipitation_group == "Wet"),
  panel.id = pan_id,
  vcov = ~GID_1
)


m_signed_ylo <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(income_group == "Poor"),
  panel.id = pan_id,
  vcov = ~GID_1
)


m_signed_ymhi <- fixest::feols(
  f_bins_signed,
  data = data_signed_groups %>% filter(income_group == "Rich"),
  panel.id = pan_id,
  vcov = ~GID_1
)

make_common_base_group_formula <- function(group_var) {
  as.formula(
    paste(
      outcome,
      "~",
      base_climate,
      paste0("+ i(TM_bin_signed, ", group_var, ", ref = 0)"),
      paste0("+ i(RR_bin_signed, ", group_var, ", ref = 0)"),
      "|",
      fixed_effects
    )
  )
}

m_signed_t_common_base <- fixest::feols(
  make_common_base_group_formula("mean_temperature_group"),
  data = data_signed_groups,
  panel.id = pan_id,
  vcov = ~GID_1
)

m_signed_p_common_base <- fixest::feols(
  make_common_base_group_formula("mean_precipitation_group"),
  data = data_signed_groups,
  panel.id = pan_id,
  vcov = ~GID_1
)

m_signed_income_common_base <- fixest::feols(
  make_common_base_group_formula("income_group"),
  data = data_signed_groups,
  panel.id = pan_id,
  vcov = ~GID_1
)

etable(
  m_bins_signed,
  m_signed_tlo,
  m_signed_thi,
  m_signed_plo,
  m_signed_phi,
  m_signed_ylo,
  m_signed_ymhi,
  headers = c("All", "Cold", "Hot", "Dry", "Wet","Poor","Rich"),
  fitstat = ~n + r2 + wr2
)

etable(
  m_signed_t_common_base,
  m_signed_p_common_base,
  m_signed_income_common_base,
  headers = c("Mean temperature", "Mean precipitation", "Income"),
  fitstat = ~n + r2 + wr2
)

extract_mean_temperature_bins <- function(model) {
  if (!inherits(model, "fixest")) {
    stop("model must be a fixest object.")
  }

  bin_var <- "TM_bin_signed"
  group_var <- "mean_temperature_group"
  interaction_marker <- paste0(":", group_var, "::")
  model_coef <- coef(model)
  coef_names <- names(model_coef)
  selected <- startsWith(coef_names, paste0(bin_var, "::")) &
    grepl(interaction_marker, coef_names, fixed = TRUE)

  if (!any(selected)) {
    stop("No temperature-bin by mean-temperature-group coefficients found.")
  }

  coef_names <- coef_names[selected]
  coefficient_tail <- substring(coef_names, nchar(bin_var) + 3L)
  model_ci <- confint(model)

  estimates <- data.frame(
    estimate = unname(model_coef[coef_names]),
    ci_low = model_ci[coef_names, 1],
    ci_high = model_ci[coef_names, 2],
    bin_label = sub(
      paste0(interaction_marker, ".*$"),
      "",
      coefficient_tail
    ),
    temperature_group = sub(
      paste0("^.*", interaction_marker),
      "",
      coefficient_tail
    )
  )

  reference <- data.frame(
    estimate = 0,
    ci_low = 0,
    ci_high = 0,
    bin_label = "0",
    temperature_group = levels(data_signed_groups$mean_temperature_group)
  )

  bind_rows(estimates, reference)
}

mean_temperature_bin_data <- extract_mean_temperature_bins(
  m_signed_t_common_base
)
bin_numeric <- suppressWarnings(
  as.numeric(mean_temperature_bin_data$bin_label)
)

if (anyNA(bin_numeric)) {
  stop("Signed temperature-bin coefficient labels must be numeric.")
}

mean_temperature_bin_data <- mean_temperature_bin_data %>%
  mutate(
    bin_label = factor(
      as.character(bin_numeric),
      levels = as.character(sort(unique(bin_numeric)))
    ),
    temperature_group = factor(
      temperature_group,
      levels = c("Cold", "Hot")
    )
  )

temperature_group_colors <- c(Cold = "#0072B2", Hot = "#D55E00")
temperature_group_shapes <- c(Cold = 16, Hot = 17)
temperature_group_dodge <- position_dodge(width = 0.65)

signed_temperature_bins_by_mean_temperature <- ggplot(
  mean_temperature_bin_data,
  aes(
    x = bin_label,
    y = estimate,
    color = temperature_group,
    shape = temperature_group
  )
) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
  geom_errorbar(
    aes(ymin = ci_low, ymax = ci_high),
    width = 0.06,
    linewidth = 0.5,
    position = temperature_group_dodge
  ) +
  geom_point(size = 2, position = temperature_group_dodge) +
  scale_color_manual(values = temperature_group_colors, drop = FALSE) +
  scale_shape_manual(values = temperature_group_shapes, drop = FALSE) +
  labs(
    title = "Signed temperature-bin coefficients by mean temperature",
    x = "Signed temperature anomaly bin (standard deviations)",
    y = "Coefficient estimate",
    color = "Country group",
    shape = "Country group"
  ) +
  theme_classic() +
  theme(legend.position = "right")

signed_temperature_bins_by_mean_temperature

iplot(
  m_signed_thi,
  m_signed_tlo,
  i.select = 1,
  main = "Signed temperature bins by mean temperature"
)
iplot(
  m_signed_phi,
  m_signed_plo,
  i.select = 2,
  main = "Signed precipitation bins by mean precipitation"
)


iplot(
  m_signed_ymhi,
  m_signed_ylo,
  i.select = 1,
  main = "Signed temperature bins by income"
)


iplot(
  m_signed_ymhi,
  m_signed_ylo,
  i.select = 2,
  main = "Signed precipitation bins by income"
)

```

# Test: Signed temperature and precipitation bins by climate zone

`mean_kg` takes the integer values 1--5 and is constant over time within every
region with nonmissing data in the source file. Notable entries reveal the five broad
Koppen--Geiger climate families: Indonesia and Singapore are coded 1; Saudi
Arabia, Egypt, and the United Arab Emirates are coded 2; France and the United
Kingdom are coded 3; Mongolia and most Russian regions are coded 4; and
Greenland is coded 5. The resulting interpretation is tropical, arid,
temperate, continental, and polar, respectively. Because the polar class has
too few estimation observations, source code 5 is merged into code 4 below and
the combined regression group is labeled Continental/Polar.

```{r load-koppen-geiger-zones}
kg_path <- file.path("econometrics", "data", "kg_data.csv")
if (!file.exists(kg_path)) {
  kg_path <- file.path("data", "kg_data.csv")
}
if (!file.exists(kg_path)) {
  stop("Cannot find econometrics/data/kg_data.csv.")
}

kg_zone_key <- data.frame(
  mean_kg = 1:4,
  climate_zone = c(
    "1 Tropical (A)",
    "2 Arid (B)",
    "3 Temperate (C)",
    "4 Continental/Polar (D/E)"
  )
)

kg_data <- read.csv(
  kg_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
) |>  select(-1) %>%
  transmute(
    GID_0 = as.character(GID_0),
    GID_1 = as.character(GID_1),
    year = as.integer(year),
    mean_kg_source = as.integer(mean_kg),
    mean_kg = if_else(mean_kg_source == 5L, 4L, mean_kg_source)
  ) %>%
  left_join(kg_zone_key, by = "mean_kg") %>%
  mutate(
    climate_zone = factor(
      climate_zone,
      levels = kg_zone_key$climate_zone
    )
  )

stopifnot(
  !anyDuplicated(kg_data[c("GID_0", "GID_1", "year")]),
  all(na.omit(unique(kg_data$mean_kg_source)) %in% 1:5),
  all(na.omit(unique(kg_data$mean_kg)) %in% kg_zone_key$mean_kg)
)

kg_zone_key

data_kg <- data %>%
  left_join(
    kg_data %>%
      select(GID_0, GID_1, year, mean_kg_source, mean_kg, climate_zone),
    by = c("GID_0", "GID_1", "year")
  )

kg_match_summary <- data_kg %>%
  filter(!is.na(.data[[outcome]])) %>%
  summarise(
    observations = n(),
    matched = sum(!is.na(climate_zone)),
    match_share = matched / observations
  )

kg_zone_sample <- data_kg %>%
  filter(!is.na(.data[[outcome]]), !is.na(climate_zone)) %>%
  group_by(climate_zone) %>%
  summarise(
    observations = n(),
    regions = n_distinct(GID_1),
    countries = n_distinct(GID_0),
    .groups = "drop"
  )

kg_match_summary
kg_zone_sample
```

Each climate-zone regression jointly estimates the signed temperature and
precipitation bins using the same controls, fixed effects, and region-clustered
standard errors as the pooled signed-bin model.

```{r test-signed-bins-by-climate-zone, fig.width=10, fig.height=6}
f_bins_signed_zone <- make_formula(paste(
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)"
))

m_bins_signed_by_climate <- fixest::feols(
  f_bins_signed_zone,
  data = data_kg %>% filter(!is.na(climate_zone)),
  panel.id = pan_id,
  vcov = ~GID_1,
  split = ~climate_zone
)

m_bins_signed_by_climate_with_i <- feols(
  dlgrp_pc_usd ~
    i(TM_bin_signed, climate_zone, ref = 0) +
    i(RR_bin_signed, climate_zone, ref = 0)
  | year +
    GID_1 +
    GID_1[year],
  data = data_kg %>% filter(!is.na(climate_zone)),
  panel.id = pan_id,
  vcov = ~GID_1
)

etable(
  m_bins_signed_by_climate,
  m_bins_signed_by_climate_with_i,
  fitstat = ~n + r2 + wr2
)

climate_zone_colors <- c(
  "#009E73", # Tropical
  "#E69F00", # Arid
  "#0072B2", # Temperate
  "#CC79A7"  # Continental/Polar
)
climate_zone_pch <- c(20, 17, 15, 18)
climate_zone_legend <- sub(
  "^[0-9]+ ",
  "",
  levels(data_kg$climate_zone)
)
names(climate_zone_colors) <- climate_zone_legend
names(climate_zone_pch) <- climate_zone_legend

extract_climate_zone_bins <- function(models, i_select) {
  if (!i_select %in% 1:2) {
    stop("i_select must be 1 (temperature) or 2 (precipitation).")
  }

  if (inherits(models, "fixest_multi")) {
    plot_parameters <- fixest::iplot(
      models,
      i.select = i_select,
      only.params = TRUE
    )
    model_tree <- attr(models, "tree")
    model_labels <- sub("^[0-9]+ ", "", model_tree$sample)

    return(
      plot_parameters$prms %>%
        transmute(
          estimate,
          ci_low,
          ci_high,
          bin_label = as.character(estimate_names),
          climate_zone = model_labels[match(id, model_tree$id)]
        )
    )
  }

  if (!inherits(models, "fixest")) {
    stop("models must be a fixest or fixest_multi object.")
  }

  bin_var <- c("TM_bin_signed", "RR_bin_signed")[[i_select]]
  model_coef <- coef(models)
  coef_names <- names(model_coef)
  selected <- startsWith(coef_names, paste0(bin_var, "::")) &
    grepl(":climate_zone::", coef_names, fixed = TRUE)

  if (!any(selected)) {
    stop(
      "No ", bin_var,
      " by climate_zone coefficients were found in the model."
    )
  }

  coef_names <- coef_names[selected]
  coefficient_tail <- substring(
    coef_names,
    nchar(bin_var) + 3L
  )
  model_ci <- confint(models)

  estimates <- data.frame(
    estimate = unname(model_coef[coef_names]),
    ci_low = model_ci[coef_names, 1],
    ci_high = model_ci[coef_names, 2],
    bin_label = sub(":climate_zone::.*$", "", coefficient_tail),
    climate_zone = sub(
      "^.*:climate_zone::",
      "",
      coefficient_tail
    )
  )

  reference <- data.frame(
    estimate = 0,
    ci_low = 0,
    ci_high = 0,
    bin_label = "0",
    climate_zone = climate_zone_legend
  )

  bind_rows(estimates, reference)
}

plot_climate_zone_bins <- function(models, i_select, title) {
  plot_data <- extract_climate_zone_bins(models, i_select)
  bin_numeric <- suppressWarnings(as.numeric(plot_data$bin_label))

  if (anyNA(bin_numeric)) {
    stop("Signed-bin coefficient labels must be numeric.")
  }

  bin_levels <- as.character(sort(unique(bin_numeric)))
  plot_data <- plot_data %>%
    mutate(
      bin_label = factor(
        as.character(bin_numeric),
        levels = bin_levels
      ),
      climate_zone = factor(
        sub("^[0-9]+ ", "", climate_zone),
        levels = climate_zone_legend
      )
    )

  bin_dodge <- position_dodge(width = 0.65)

  ggplot(
    plot_data,
    aes(
      x = bin_label,
      y = estimate,
      color = climate_zone,
      shape = climate_zone
    )
  ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    geom_errorbar(
      aes(ymin = ci_low, ymax = ci_high),
      width = 0.06,
      linewidth = 0.5,
      position = bin_dodge
    ) +
    geom_point(size = 2, position = bin_dodge) +
    scale_color_manual(values = climate_zone_colors, drop = FALSE) +
    scale_shape_manual(values = climate_zone_pch, drop = FALSE) +
    labs(
      title = title,
      x = "Signed anomaly bin (standard deviations)",
      y = "Coefficient estimate",
      color = "Climate zone",
      shape = "Climate zone"
    ) +
    facet_wrap(climate_zone~.,) +
    theme_classic() +
    theme(legend.position = "right") +
    coord_cartesian(ylim = c(-0.1, 0.05))
}

plot_climate_zone_bins(
  m_bins_signed_by_climate,
  i_select = 1,
  title = paste(
    "Signed temperature bins by climate zone",
    "(shared climate coefficients)"
  )
)

plot_climate_zone_bins(
  m_bins_signed_by_climate,
  i_select = 2,
  title = paste(
    "Signed precipitation bins by climate zone",
    "(shared climate coefficients)"
  )
)
```

# Test 7: Heterogeneity by climate variability

These split-sample regressions test whether the baseline deviation response
differs between regions below and above the median long-run variability of
precipitation or temperature.

```{r test-climate-variability-splits}
mplow <- fixest::feols(
  f,
  data = data %>% filter(sd_RR_all < median(sd_RR_all, na.rm = TRUE)),
  panel.id = pan_id
)
mphi <- fixest::feols(
  f,
  data = data %>% filter(sd_RR_all > median(sd_RR_all, na.rm = TRUE)),
  panel.id = pan_id
)

mtlow <- fixest::feols(
  f,
  data = data %>% filter(sd_TM_all < median(sd_TM_all, na.rm = TRUE)),
  panel.id = pan_id
)
mthi <- fixest::feols(
  f,
  data = data %>% filter(sd_TM_all > median(sd_TM_all, na.rm = TRUE)),
  panel.id = pan_id
)

etable(
  mplow,
  mphi,
  mtlow,
  mthi,
  headers = c("Low RR SD", "High RR SD", "Low TM SD", "High TM SD"),
  fitstat = ~n + r2 + wr2
)
```

# Test 8: Heterogeneity by average income

The final test compares the baseline response in the bottom and top quartiles of
region-level average GDP per capita.

```{r test-income-splits}
mylow <- fixest::feols(
  f,
  data = data_income %>% filter(av_y < income_cutoffs[[2]]),
  panel.id = pan_id
)
myhi <- fixest::feols(
  f,
  data = data_income %>% filter(av_y >= income_cutoffs[[2]]),
  panel.id = pan_id
)

etable(
  mylow,
  myhi,
  headers = c("Bottom income quartile", "Top income quartile"),
  fitstat = ~n + r2 + wr2
)
```

# Test 9: Asymmetric temperature--precipitation interactions

This specification allows positive and negative temperature and precipitation
deviations to have separate effects. The four interactions additionally allow
the response to differ across the hot--wet, hot--dry, cold--wet, and cold--dry
quadrants. Each component is expressed as a non-negative anomaly magnitude.

```{r test-asymmetric-interactions}


model_asymmetric_interactions <- paste(
  base_climate,
  "+ abs_zTMp^2 + abs_zTMm^2 + abs_zRRp + abs_zRRm",
  "+ abs_zTMp^2:abs_zRRp + abs_zTMp^2:abs_zRRm",
  "+ abs_zTMm^2:abs_zRRp + abs_zTMm^2:abs_zRRm"
)

f_asymmetric_interactions <- make_formula(model_asymmetric_interactions)
m_asymmetric_interactions <- fit_climate_model(model_asymmetric_interactions)

m_asym_low <- fixest::feols(
  f_asymmetric_interactions,
  data = data_income %>% filter(av_y < income_cutoffs[[2]]),
  panel.id = pan_id,
    cluster = ~GID_1
)
m_asym_hi <- fixest::feols(
  f_asymmetric_interactions,
  data = data_income %>% filter(av_y >= income_cutoffs[[2]]),
  panel.id = pan_id, 
    cluster = ~GID_1
)
etable(
  m_asymmetric_interactions,
  m_asym_low,
  m_asym_hi,
  fitstat = ~n + r2 + wr2
)
```

# Test 10: Climate velocity during warming and cooling runs

Here, `climate_velocity` is a short-run temporal warming rate rather than a
spatial velocity. For year $t$, it is estimated from the fixed trailing
`r params$climate_velocity_years`-year window ending in $t$. The window therefore
uses no future observations. The slope is expressed in degrees Celsius per
decade and is calculated independently of whether year $t$ belongs to a warming
run.

A warming run is a maximal sequence of at least two consecutive calendar years
with strictly increasing annual temperature. Position 1 is the year immediately
before the first positive change, and position 2 is the first warmer year.
`warming_run_rate` records the separate slope
estimated only from a run's observations. `warming_run_length` records the
year's position within its run, while `warming_run_total_length` preserves the
run's eventual complete length. Single-year events are coded as 1 rather than
0, making them equivalent to the first year of a subsequent run. Cooling runs
and their corresponding fields are defined symmetrically from strictly
decreasing annual temperature. A local peak or trough can therefore be the
final year of one run and the first year of the opposite run.
`cumulative_warming` and `cumulative_cooling` sum the annual temperature
changes belonging to the corresponding run through the current year. They are
zero in position 1; cumulative warming is positive and cumulative cooling is
negative thereafter.

The fixed-baseline deviations hold the lagged 20-year temperature mean and
standard deviation at their position-1 values throughout each run. Because a
turning point belongs to both an ending run and a new run, the standardized
deviations are retained separately as `zTM_warming_f` and `zTM_cooling_f`.
Their one-sided magnitudes are `zTMp_f` and `zTMm_f`. Outside the corresponding
run, and at position 1, these equal the original moving-baseline deviations.

```{r construct-climate-velocity}
fit_warming_run_rate <- function(year, temperature) {
  stopifnot(
    length(year) == length(temperature),
    length(year) >= 2L,
    all(is.finite(year)),
    all(is.finite(temperature)),
    all(diff(year) == 1L),
    all(diff(temperature) > 0)
  )

  # The fitted annual slope is multiplied by 10 to obtain degrees per decade.
  unname(stats::lm.fit(
    x = cbind(1, year),
    y = temperature
  )$coefficients[[2]]) * 10
}

fit_cooling_run_rate <- function(year, temperature) {
  stopifnot(
    length(year) == length(temperature),
    length(year) >= 2L,
    all(is.finite(year)),
    all(is.finite(temperature)),
    all(diff(year) == 1L),
    all(diff(temperature) < 0)
  )

  # Cooling slopes remain negative and are expressed in degrees per decade.
  unname(stats::lm.fit(
    x = cbind(1, year),
    y = temperature
  )$coefficients[[2]]) * 10
}

fit_trailing_climate_velocity <- function(year, temperature, window_years) {
  stopifnot(
    length(year) == length(temperature),
    length(window_years) == 1L,
    window_years >= 3L
  )

  vapply(seq_along(year), function(i) {
    first <- i - window_years + 1L

    if (first < 1L) {
      return(NA_real_)
    }

    idx <- seq.int(first, i)
    year_window <- year[idx]
    temperature_window <- temperature[idx]

    if (any(!is.finite(year_window)) ||
        any(!is.finite(temperature_window)) ||
        !all(diff(year_window) == 1L)) {
      return(NA_real_)
    }

    # A trailing annual slope, converted from degrees/year to degrees/decade.
    unname(stats::lm.fit(
      x = cbind(1, year_window),
      y = temperature_window
    )$coefficients[[2]]) * 10
  }, numeric(1))
}

climate_velocity_hist <- data %>%
  group_by(GID_1) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    consecutive_year = year == dplyr::lag(year) + 1L,
    dTM_run = if_else(consecutive_year, TM - dplyr::lag(TM), NA_real_),
    warming_change = !is.na(dTM_run) & dTM_run > 0,
    warming_run_start = warming_change &
      !dplyr::lag(warming_change, default = FALSE),
    warming_change_run_id = if_else(
      warming_change,
      cumsum(warming_run_start),
      NA_integer_
    ),
    # Attach the year preceding the first increase to the same run.
    warming_run_id = coalesce(
      warming_change_run_id,
      if_else(
        dplyr::lead(warming_change, default = FALSE),
        dplyr::lead(warming_change_run_id),
        NA_integer_
      )
    ),
    cooling_change = !is.na(dTM_run) & dTM_run < 0,
    cooling_run_start = cooling_change &
      !dplyr::lag(cooling_change, default = FALSE),
    cooling_change_run_id = if_else(
      cooling_change,
      cumsum(cooling_run_start),
      NA_integer_
    ),
    # Attach the year preceding the first decrease to the same run.
    cooling_run_id = coalesce(
      cooling_change_run_id,
      if_else(
        dplyr::lead(cooling_change, default = FALSE),
        dplyr::lead(cooling_change_run_id),
        NA_integer_
      )
    ),
    climate_velocity = fit_trailing_climate_velocity(
      year,
      TM,
      window_years = CLIMATE_VELOCITY_YEARS
    )
  ) %>%
  group_by(GID_1, warming_run_id) %>%
  mutate(
    warming_run = !is.na(warming_run_id) & n() >= 2L,
    warming_run_length = if_else(warming_run, row_number(), 1L),
    warming_run_total_length = if_else(warming_run, n(), 1L),
    # Sum within-run dTM through this year; position 1 has no within-run change.
    cumulative_warming = if_else(warming_run, TM - first(TM), 0),
    mean_TM_warming_f = if_else(
      warming_run,
      first(mean_TM_lag_30),
      mean_TM_lag_30
    ),
    sd_TM_warming_f = if_else(
      warming_run,
      first(sd_TM_lag_30),
      sd_TM_lag_30
    ),
    zTM_warming_f = (TM - mean_TM_warming_f) / sd_TM_warming_f,
    warming_run_rate = if (!first(warming_run)) {
      NA_real_
    } else {
      fit_warming_run_rate(year, TM)
    },
    warming_run_id = if_else(warming_run, warming_run_id, NA_integer_)
  ) %>%
  ungroup() %>%
  group_by(GID_1, cooling_run_id) %>%
  mutate(
    cooling_run = !is.na(cooling_run_id) & n() >= 2L,
    cooling_run_length = if_else(cooling_run, row_number(), 1L),
    cooling_run_total_length = if_else(cooling_run, n(), 1L),
    # Cooling dTM values retain their negative sign.
    cumulative_cooling = if_else(cooling_run, TM - first(TM), 0),
    mean_TM_cooling_f = if_else(
      cooling_run,
      first(mean_TM_lag_30),
      mean_TM_lag_30
    ),
    sd_TM_cooling_f = if_else(
      cooling_run,
      first(sd_TM_lag_30),
      sd_TM_lag_30
    ),
    zTM_cooling_f = (TM - mean_TM_cooling_f) / sd_TM_cooling_f,
    cooling_run_rate = if (!first(cooling_run)) {
      NA_real_
    } else {
      fit_cooling_run_rate(year, TM)
    },
    cooling_run_id = if_else(cooling_run, cooling_run_id, NA_integer_)
  ) %>%
  ungroup() %>%
  mutate(
    # Activate velocity only during the corresponding runs without discarding
    # observations outside those runs.
    climate_velocity_warming = if_else(
      warming_run & climate_velocity>0,
      climate_velocity,
      0
    ),
    climate_velocity_cooling = if_else(
      cooling_run & climate_velocity<0,
      climate_velocity,
      0
    ),
    zTMp_f = pmax(zTM_warming_f, 0),
    zTMm_f = pmax(-zTM_cooling_f, 0),
    # Elapsed run position, with the fifth and later years grouped as 5+.
    warming_run_length_capped = pmin(warming_run_length, 5L),
    cooling_run_length_capped = pmin(cooling_run_length, 5L)
  )

# Every run includes its first year and contains at least two observations.
stopifnot(all(climate_velocity_hist$warming_run_length >= 1L))
stopifnot(all(climate_velocity_hist$cooling_run_length >= 1L))
stopifnot(all(climate_velocity_hist$cumulative_warming >= 0))
stopifnot(all(climate_velocity_hist$cumulative_cooling <= 0))
stopifnot(isTRUE(all.equal(
  climate_velocity_hist$zTMp_f[
    climate_velocity_hist$warming_run_length == 1L
  ],
  climate_velocity_hist$abs_zTMp[
    climate_velocity_hist$warming_run_length == 1L
  ],
  check.attributes = FALSE
)))
stopifnot(isTRUE(all.equal(
  climate_velocity_hist$zTMm_f[
    climate_velocity_hist$cooling_run_length == 1L
  ],
  climate_velocity_hist$abs_zTMm[
    climate_velocity_hist$cooling_run_length == 1L
  ],
  check.attributes = FALSE
)))
stopifnot(all(
  climate_velocity_hist$cumulative_warming[
    climate_velocity_hist$warming_run_length == 1L
  ] == 0
))
stopifnot(all(
  climate_velocity_hist$cumulative_cooling[
    climate_velocity_hist$cooling_run_length == 1L
  ] == 0
))
stopifnot(all(
  climate_velocity_hist$warming_run |
    climate_velocity_hist$cooling_run
))
stopifnot(all(
  !climate_velocity_hist$warming_run |
    climate_velocity_hist$warming_run_total_length >= 2L
))
stopifnot(all(
  !climate_velocity_hist$warming_run |
    climate_velocity_hist$warming_run_length <=
      climate_velocity_hist$warming_run_total_length
))
stopifnot(all(
  !climate_velocity_hist$cooling_run |
    climate_velocity_hist$cooling_run_total_length >= 2L
))
stopifnot(all(
  !climate_velocity_hist$cooling_run |
    climate_velocity_hist$cooling_run_length <=
      climate_velocity_hist$cooling_run_total_length
))

warming_run_summary <- climate_velocity_hist %>%
  filter(warming_run) %>%
  group_by(GID_1, warming_run_id) %>%
  summarise(
    warming_run_total_length = first(warming_run_total_length),
    warming_run_rate = first(warming_run_rate),
    .groups = "drop"
  ) %>%
  arrange(GID_1, warming_run_id)

warming_run_summary %>%
  count(warming_run_total_length, name = "runs")

cooling_run_summary <- climate_velocity_hist %>%
  filter(cooling_run) %>%
  group_by(GID_1, cooling_run_id) %>%
  summarise(
    cooling_run_total_length = first(cooling_run_total_length),
    cooling_run_rate = first(cooling_run_rate),
    .groups = "drop"
  ) %>%
  arrange(GID_1, cooling_run_id)

cooling_run_summary %>%
  count(cooling_run_total_length, name = "runs")

climate_velocity_hist %>%
  filter(warming_run, !is.na(climate_velocity)) %>%
  ggplot(aes(
    x = climate_velocity_warming,
    color = factor(warming_run_total_length)
  )) +
  geom_density() +
  labs(color = "Warming-run length")

climate_velocity_hist %>%
  filter(cooling_run, !is.na(climate_velocity)) %>%
  ggplot(aes(
    x = climate_velocity_cooling,
    color = factor(cooling_run_total_length)
  )) +
  geom_density() +
  labs(color = "Cooling-run length")
 
climate_velocity_hist |> filter(warming_run_total_length>=7) |> group_by(GID_1) |> filter(year==min(year)) |>  ungroup() |> select(GID_0,year) |> unique()


climate_velocity_hist |> filter(warming_run_total_length>=6) |> group_by(GID_1) |> filter(year==max(year)) |>  ungroup() |> select(GID_1,year,cumulative_warming) |> unique() |> 
  ggplot() + geom_density(aes(x=cumulative_warming))
```

## regressions for climate velocity

```{r test-climate-velocity}

model_velocity_bool <- paste(
  base_climate, 
  "+ zRR^2 +",
  "+ I(zTMp_f^2) : i(warming_run_length)",
  "+ I(zTMm_f^2) : i(cooling_run_length)",
  "+ l(zTMp_f^2,1:3) + l(zTMm_f^2,1:3)"
)

m_bool <- fit_climate_model(
  model_velocity_bool,
  model_data = climate_velocity_hist
)



model_cumulative_change <- paste(
  base_climate,
  "+ zRR^2 +",
  "+ I(zTMp_f^2) : cumulative_warming",
  "+ I(zTMp_f^2) : cumulative_cooling",
  "+ I(zTMp_f^2) + I(zTMp_f^2)",
  "+ l(zTMp_f,1:3) + l(zTMm_f,1:3)"
)

m_cumulative_change <- fit_climate_model(
  model_cumulative_change,
  model_data = climate_velocity_hist
)

bin_nonzero_magnitude <- function(x, n_bins = 5L) {
  if (any(x < 0, na.rm = TRUE)) {
    stop("Binned magnitudes must be non-negative.")
  }

  positive <- !is.na(x) & x > 0
  result <- rep(NA_character_, length(x))
  result[!is.na(x) & x == 0] <- "0"

  if (!any(positive)) {
    return(factor(result, levels = "0"))
  }

  positive_bins <- fixest::bin(
    x[positive],
    paste0("cut::", n_bins)
  )
  result[positive] <- as.character(positive_bins)

  factor(result, levels = c("0", levels(positive_bins)))
}

climate_velocity_hist <- climate_velocity_hist %>%
  mutate(
    climate_velocity_warming_bin = bin_nonzero_magnitude(
      climate_velocity_warming
    ),
    climate_velocity_cooling_bin = bin_nonzero_magnitude(
      abs(climate_velocity_cooling)
    ),
    cumulative_warming_bin = bin_nonzero_magnitude(
      cumulative_warming
    ),
    cumulative_cooling_bin = bin_nonzero_magnitude(
      abs(cumulative_cooling)
    )
  )



model_velocity_bin <- paste(
  base_climate,
  paste0(" + i(climate_velocity_warming_bin, I(zTMp_f^2))",
  "+ l(zTMp_f^2,1:3) + l(zTMm_f^2,0:3)"
  )
)

f_velocity <- make_formula(model_velocity_bin)
m_velocity <- fit_climate_model(
  model_velocity_bin,
  model_data = climate_velocity_hist
)

model_cumulative_bin <- paste(
  base_climate,
  paste0(
    "+  ",
    "i(cumulative_warming_bin, I(zTMp_f^2))"
  ),
  paste0(
    "+  ",
    "i(cumulative_cooling_bin, I(zTMm_f^2))",
  "+ l(zTMp_f^2,1:3) + l(zTMm_f^2,1:3)"
  )
)

f_cumulative_bin <- make_formula(model_cumulative_bin)
m_cumulative_bin <- fit_climate_model(
  model_cumulative_bin,
  model_data = climate_velocity_hist
)

etable(
  m_velocity,
  m_cumulative_bin,
  m_bool,
  m_cumulative_change,
  fitstat = ~n + r2 + wr2
)
iplot(
  m_velocity, 
  main = "Temperature response and trailing climate velocity"
)

iplot(
  m_cumulative_bin, i.select=1,
  main = "Temperature response and cumulative run change (warming)"
)

iplot(
  m_cumulative_bin, i.select=2,
  main = "Temperature response and cumulative run change (cooling)"
)

iplot(
  m_bool, 
  main = "Temperature response and consecutive warming years"
)

etable(m_bool, fitstat = ~n + r2 + wr2)

```

## Association between cumulative-change and climate-velocity bins

The association is calculated separately for warming and cooling. Spearman's
rho treats the bins as ordered, while Cramer's V treats them as categorical.
The jointly non-zero results isolate active cumulative change and climate
velocity from their shared zero reference category.

```{r cumulative-velocity-bin-association, fig.width=10, fig.height=5}
summarise_bin_association <- function(
    data,
    cumulative_bin,
    velocity_bin,
    direction,
    scope = c("All complete", "Both non-zero")) {
  scope <- match.arg(scope)

  pairs <- data %>%
    transmute(
      cumulative = .data[[cumulative_bin]],
      velocity = .data[[velocity_bin]]
    ) %>%
    filter(!is.na(cumulative), !is.na(velocity))

  if (scope == "Both non-zero") {
    pairs <- pairs %>%
      filter(as.character(cumulative) != "0", as.character(velocity) != "0")
  }

  contingency <- table(
    droplevels(pairs$cumulative),
    droplevels(pairs$velocity)
  )
  enough_bins <- nrow(contingency) >= 2L && ncol(contingency) >= 2L

  spearman_rho <- if (enough_bins) {
    stats::cor(
      as.integer(pairs$cumulative),
      as.integer(pairs$velocity),
      method = "spearman"
    )
  } else {
    NA_real_
  }

  cramers_v <- if (enough_bins) {
    chi_squared <- suppressWarnings(stats::chisq.test(
      contingency,
      correct = FALSE
    ))$statistic
    sqrt(
      unname(chi_squared) /
        (sum(contingency) * min(
          nrow(contingency) - 1L,
          ncol(contingency) - 1L
        ))
    )
  } else {
    NA_real_
  }

  tibble::tibble(
    direction = direction,
    scope = scope,
    observations = nrow(pairs),
    spearman_rho = spearman_rho,
    cramers_v = cramers_v
  )
}

bin_association_summary <- dplyr::bind_rows(
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_warming_bin",
    "climate_velocity_warming_bin",
    "Warming",
    "All complete"
  ),
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_warming_bin",
    "climate_velocity_warming_bin",
    "Warming",
    "Both non-zero"
  ),
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_cooling_bin",
    "climate_velocity_cooling_bin",
    "Cooling",
    "All complete"
  ),
  summarise_bin_association(
    climate_velocity_hist,
    "cumulative_cooling_bin",
    "climate_velocity_cooling_bin",
    "Cooling",
    "Both non-zero"
  )
)

bin_association_summary

joint_active_bins <- dplyr::bind_rows(
  climate_velocity_hist %>%
    transmute(
      direction = "Warming",
      cumulative_bin = cumulative_warming_bin,
      velocity_bin = climate_velocity_warming_bin
    ),
  climate_velocity_hist %>%
    transmute(
      direction = "Cooling",
      cumulative_bin = cumulative_cooling_bin,
      velocity_bin = climate_velocity_cooling_bin
    )
) %>%
  filter(
    !is.na(cumulative_bin),
    !is.na(velocity_bin),
    as.character(cumulative_bin) != "0",
    as.character(velocity_bin) != "0"
  ) %>%
  count(direction, cumulative_bin, velocity_bin, name = "observations") %>%
  group_by(direction) %>%
  mutate(share = observations / sum(observations)) %>%
  ungroup()

ggplot(
  joint_active_bins,
  aes(x = velocity_bin, y = cumulative_bin, fill = share)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", 100 * share)), size = 3) +
  facet_wrap(~direction, scales = "free") +
  labs(
    x = "Climate-velocity magnitude bin",
    y = "Cumulative-change magnitude bin",
    fill = "Share",
    title = "Joint distribution among non-zero run indicators"
  ) +
  theme_minimal()
```

## test only lags that belong to the same warming run

```{r}


add_same_run_lags <- function(
    data,
    value_column,
    run_length_column,
    prefix,
    max_lag) {
  data <- data %>%
    arrange(GID_1, year) %>%
    group_by(GID_1)

  for (lag_number in seq_len(max_lag)) {
    lag_column <- paste0(prefix, lag_number, "_same")
    data <- data %>%
      mutate(
        !!lag_column := if_else(
          .data[[run_length_column]] > lag_number,
          dplyr::lag(.data[[value_column]], n = lag_number),
          0
        )
      )
  }

  ungroup(data)
}

climate_velocity_hist <- climate_velocity_hist %>%
  arrange(GID_1, year) %>%
  mutate(
    zTMp2 = zTMp_f^2,
    zTMm2 = zTMm_f^2
  )

# Generate every feasible lag rather than imposing an arbitrary three-year cap.
max_warming_same_run_lag <- max(
  climate_velocity_hist$warming_run_length,
  na.rm = TRUE
) - 1L
max_cooling_same_run_lag <- max(
  climate_velocity_hist$cooling_run_length,
  na.rm = TRUE
) - 1L

climate_velocity_hist <- add_same_run_lags(
  climate_velocity_hist,
  value_column = "zTMp2",
  run_length_column = "warming_run_length",
  prefix = "zTMp2_l",
  max_lag = max_warming_same_run_lag
) %>%
  add_same_run_lags(
    value_column = "zTMm2",
    run_length_column = "cooling_run_length",
    prefix = "zTMm2_l",
    max_lag = max_cooling_same_run_lag
  )

warming_same_run_lag_terms <- paste0(
  "zTMp2_l",
  seq_len(max_warming_same_run_lag),
  "_same"
)
cooling_same_run_lag_terms <- paste0(
  "zTMm2_l",
  seq_len(max_cooling_same_run_lag),
  "_same"
)
same_run_lag_terms <- c(
  warming_same_run_lag_terms,
  cooling_same_run_lag_terms
)
same_run_lag_formula <- if (length(same_run_lag_terms)) {
  paste("+", paste(same_run_lag_terms, collapse = " + "))
} else {
  ""
}

# A lag is zero whenever looking back that far would leave the current run.
for (lag_number in seq_len(max_warming_same_run_lag)) {
  lag_column <- warming_same_run_lag_terms[[lag_number]]
  stopifnot(all(
    climate_velocity_hist[[lag_column]][
      climate_velocity_hist$warming_run_length <= lag_number
    ] == 0
  ))
}
for (lag_number in seq_len(max_cooling_same_run_lag)) {
  lag_column <- cooling_same_run_lag_terms[[lag_number]]
  stopifnot(all(
    climate_velocity_hist[[lag_column]][
      climate_velocity_hist$cooling_run_length <= lag_number
    ] == 0
  ))
}

model_velocity_bool <- paste(
  base_climate, 
  "+ abs_zRRm + abs_zRRp +",
  "+ I(zTMp_f^2) + I(zTMm_f^2)",
  same_run_lag_formula
)

```


# Test 11: the SD model vs the no SD model


```{r test SD vs no SD}

data_nosd <- data |> mutate(TM_bin_signed=TM_bin_signed*sd_TM_lag_20,
                           TM_bin_signed=TM_bin_signed*sd_TM_lag_20)
model_nosd <- paste(
  base_climate,
  paste0(
    "+ i(TM_bin_signed, ref = 0) ",
    "+ i(RR_bin_signed, ref = 0) "
  )
)

m_nosd <- fit_climate_model(
  model_nosd,
  model_data = data_nosd
)

etable(m_bins_signed, m_nosd, fitstat = ~n + r2 + wr2)
iplot(
  m_nosd, m_bins_signed, i.select=1,
  main = "Temperature response"
)
iplot(
  m_nosd, m_bins_signed, i.select=2,
  main = "Precipitation response"
)

```

```{r test adaptation}

ADAPTATION_SPLINE_DF <- 3L

data_adaptation <- data %>%
  group_by(gadm_level, GID_1) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    lag_lgrp_pc_usd_1 = lag_by_year(lgrp_pc_usd, year)
  ) %>%
  ungroup()

# `ns()` returns a matrix, which cannot be multiplied directly by the matrix
# created by `i()`. Store each spline basis as a scalar column so fixest can
# construct the bin-by-basis interactions separately.
valid_income_lag <- is.finite(data_adaptation$lag_lgrp_pc_usd_1)
income_spline <- matrix(
  NA_real_,
  nrow = nrow(data_adaptation),
  ncol = ADAPTATION_SPLINE_DF
)
income_spline_fit <- splines::ns(
  data_adaptation$lag_lgrp_pc_usd_1[valid_income_lag],
  df = ADAPTATION_SPLINE_DF
)
income_spline[valid_income_lag, ] <- income_spline_fit
spline_columns <- paste0("lag_lgrp_spline_", seq_len(ADAPTATION_SPLINE_DF))
colnames(income_spline) <- spline_columns

data_adaptation <- bind_cols(
  data_adaptation,
  tibble::as_tibble(income_spline)
)

temperature_adaptation_terms <- paste0(
  "i(TM_bin_signed, ", spline_columns, ", ref = 0)"
)
precipitation_adaptation_terms <- paste0(
  "i(RR_bin_signed, ", spline_columns, ", ref = 0)"
)

model_ada <- paste(
  base_climate,
  "+ i(TM_bin_signed, ref = 0)",
  "+ i(RR_bin_signed, ref = 0)",
  "+", paste(
    c(temperature_adaptation_terms, precipitation_adaptation_terms),
    collapse = " + "
  )
)

m_ada <- fit_climate_model(
  model_ada,
  model_data = data_adaptation
)

etable(m_ada, fitstat = ~n + r2 + wr2)

# Evaluate the fitted response at three representative lagged-income levels.
income_probabilities <- c(0.05, 0.50, 0.95)
income_percentile_names <- c("5th percentile", "Median", "95th percentile")
income_panel_rows <- est_row(data_adaptation) & valid_income_lag
income_log_levels <- unname(stats::quantile(
  data_adaptation$lag_lgrp_pc_usd_1[income_panel_rows],
  probs = income_probabilities,
  na.rm = TRUE
))

# Use the knots and boundary knots from the spline fitted in the model data.
income_spline_levels <- predict(
  income_spline_fit,
  newx = income_log_levels
)
colnames(income_spline_levels) <- spline_columns

income_scenarios <- tibble::tibble(
  income_percentile = income_percentile_names,
  log_income = income_log_levels,
  income = exp(income_log_levels),
  income_label = paste0(
    income_percentile_names,
    " ($",
    format(
      round(exp(income_log_levels)),
      big.mark = ",",
      scientific = FALSE,
      trim = TRUE
    ),
    ")"
  )
)
income_scenarios$income_label <- factor(
  income_scenarios$income_label,
  levels = income_scenarios$income_label
)

adaptation_response_data <- function(
    fitted_model,
    bin_var,
    model_data,
    spline_values,
    scenarios
) {
  model_coef <- stats::coef(fitted_model)
  model_vcov <- stats::vcov(fitted_model)
  coefficient_names <- names(model_coef)
  bin_values <- sort(unique(model_data[[bin_var]][
    !is.na(model_data[[bin_var]])
  ]))

  purrr::map_dfr(seq_len(nrow(scenarios)), function(income_id) {
    purrr::map_dfr(bin_values, function(bin_value) {
      if (bin_value == 0) {
        return(tibble::tibble(
          bin = bin_value,
          estimate = 0,
          std_error = 0
        ))
      }

      main_coefficient <- paste0(bin_var, "::", bin_value)
      interaction_coefficients <- paste0(
        main_coefficient,
        ":",
        spline_columns
      )
      required_coefficients <- c(
        main_coefficient,
        interaction_coefficients
      )
      missing_coefficients <- setdiff(
        required_coefficients,
        coefficient_names
      )
      if (length(missing_coefficients)) {
        stop(
          "Adaptation coefficients missing from the model: ",
          paste(missing_coefficients, collapse = ", ")
        )
      }

      contrast <- setNames(
        numeric(length(model_coef)),
        coefficient_names
      )
      contrast[[main_coefficient]] <- 1
      contrast[interaction_coefficients] <- spline_values[income_id, ]

      estimate <- sum(contrast * model_coef)
      variance <- drop(t(contrast) %*% model_vcov %*% contrast)

      tibble::tibble(
        bin = bin_value,
        estimate = estimate,
        std_error = sqrt(max(variance, 0))
      )
    }) %>%
      mutate(
        income_label = scenarios$income_label[[income_id]],
        income = scenarios$income[[income_id]]
      )
  }) %>%
    mutate(
      conf_low = estimate - stats::qnorm(0.975) * std_error,
      conf_high = estimate + stats::qnorm(0.975) * std_error
    )
}

plot_adaptation_response <- function(plot_data, climate_variable) {
  response_dodge <- position_dodge(width = 0.12)

  ggplot(
    plot_data,
    aes(
      x = bin,
      y = estimate,
      color = income_label,
      group = income_label
    )
  ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    geom_line(linewidth = 0.6) +
    geom_errorbar(
      aes(ymin = conf_low, ymax = conf_high),
      width = 0.06,
      linewidth = 0.5,
      position = response_dodge
    ) +
    geom_point(size = 2, position = response_dodge) +
    scale_x_continuous(breaks = sort(unique(plot_data$bin))) +
    labs(
      title = paste(climate_variable, "response by lagged income"),
      x = "Signed anomaly bin (standard deviations)",
      y = "Estimated response",
      color = "Lagged GDP per capita"
    ) +
    theme_classic() +
    theme(legend.position = "right")
}

temperature_response_by_income <- adaptation_response_data(
  m_ada,
  bin_var = "TM_bin_signed",
  model_data = data_adaptation,
  spline_values = income_spline_levels,
  scenarios = income_scenarios
)
precipitation_response_by_income <- adaptation_response_data(
  m_ada,
  bin_var = "RR_bin_signed",
  model_data = data_adaptation,
  spline_values = income_spline_levels,
  scenarios = income_scenarios
)

plot_adaptation_response(
  temperature_response_by_income,
  "Temperature"
)
plot_adaptation_response(
  precipitation_response_by_income,
  "Precipitation"
)

```

## GDP growth during extreme temperature and precipitation shocks

The violins compare residual GDP-per-capita growth in years with shocks within
2.5 standard deviations of a region's moving climate baseline against the very
hot/wet and very cold/dry tails. Growth is first residualized using region and
year fixed effects plus region-specific linear year trends, matching the
`GID_1 + year + GID_1[year]` convention. Temperature and precipitation are
shown in separate panels because a region-year can belong to different tail
categories for the two climate variables.

```{r violin-gdp-growth-extreme-shocks, fig.width=10, fig.height=5.5}
shock_levels <- c(
  "Within +/-2.5 SD",
  "> 2.5 SD (very hot/wet)",
  "< -2.5 SD (very cold/dry)"
)

gdp_growth_fe_data <- data %>%
  ungroup() %>%
  filter(
    is.finite(dlgrp_pc_usd),
    !is.na(GID_1),
    !is.na(year)
  )

gdp_growth_fe_model <- fixest::feols(
  dlgrp_pc_usd ~ 1 | GID_1 + year + GID_1[year],
  data = gdp_growth_fe_data
)

gdp_growth_fe_data <- gdp_growth_fe_data %>%
  mutate(
    # Keep an NA placeholder for observations removed internally by fixest
    # (for example, fixed-effect singletons) so rows remain aligned.
    gdp_growth_residual = stats::residuals(
      gdp_growth_fe_model,
      na.rm = FALSE
    )
  )

gdp_growth_by_shock <- dplyr::bind_rows(
  gdp_growth_fe_data %>%
    transmute(
      climate_variable = "Temperature",
      shock = zTM,
      gdp_growth = gdp_growth_residual
    ),
  gdp_growth_fe_data %>%
    transmute(
      climate_variable = "Precipitation",
      shock = zRR,
      gdp_growth = gdp_growth_residual
    )
) %>%
  filter(is.finite(shock), is.finite(gdp_growth)) %>%
  mutate(
    shock_group = case_when(
      shock > 2.5 ~ "> 2.5 SD (very hot/wet)",
      shock < -2.5 ~ "< -2.5 SD (very cold/dry)",
      TRUE ~ "Within +/-2.5 SD"
    ),
    shock_group = factor(shock_group, levels = shock_levels),
    climate_variable = factor(
      climate_variable,
      levels = c("Temperature", "Precipitation")
    )
  )

shock_group_counts <- gdp_growth_by_shock %>%
  count(climate_variable, shock_group, .drop = FALSE)

shock_group_counts

ggplot(
  gdp_growth_by_shock,
  aes(x = shock_group, y = 100 * gdp_growth, fill = shock_group)
) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = "grey30",
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    fill = "white",
    alpha = 0.8,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ climate_variable) +
  scale_fill_manual(
    values = c(
      "Within +/-2.5 SD" = "grey70",
      "> 2.5 SD (very hot/wet)" = "#D55E00",
      "< -2.5 SD (very cold/dry)" = "#0072B2"
    ),
    drop = FALSE
  ) +
  labs(
    title = "Residual GDP growth by standardized climate shock",
    subtitle = paste(
      "Growth residualized by region and year fixed effects",
      "and region-specific linear trends"
    ),
    x = NULL,
    y = "Residual GDP per-capita growth (percentage points)",
    fill = "Shock category"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "none"
  )
```


